---
title: "Machine learning essentials"
author: "Michael Love"
date: 11/16/2018
output: html_document
---

The last module of this course will be an overview of Machine Learning
(ML), including machine learning essentials, and introduction to
support vector machines (SVM), random forests (RF), and neural
networks (NN). Here we begin with some conceptual and procedural
frameworks that are essential for conducting analysis or research with
machine learning tools. The most important considerations, which span
across a variety of tools, are:

* how models are trained / "fit"
    - feature selection
    - parameter tuning
    - iterative fitting (data splitting)
* how models are evaluated
    - training vs test set evaluation

We will spend most of the lecture note on discussing various
frameworks for how models are fit. We will first briefly cover some of
the metrics used for evaluation of predictive models. 

# Metrics for evaluation of models

After fitting a model, we can use it to predict the value of the same
data used for fitting the model, the *training set*, or on new data,
the *test set*. As statisticians, you should now be familiar with the
idea that the performance on the model on the training set will
exaggerate the models performance on new data, due to
*overfitting*. While some methods are more prone to overfitting than
others, it is an unavoidable property of model training, which we will
discuss more in this note. In either case we can evaluate the
prediction with various metrics. For continuous data, three common
evaluation metrics are:

* root mean squared error
* predictive $R^2$ (multiple definitions)
* mean absolute error

Two common definitions for predictive $R^2$ are the squared Pearson
correlation of the true values with the predicted values (which cannot
be negative) and 1 - sum of squared residuals / total sum of
squares (which can be negative). If a predictive $R^2$ using the
second metric is negative, it means the prediction was worse than just
guessing the mean value, which could indicate a bug or that the model
has a lot of variance.

For categorical data, there are also various evaluation metrics. If
the outcome is binary, it is common to consider sensitivity,
specificity, and precision. Two generic methods though are:

* accuracy - percentage of correctly classified
* [Cohen's kappa](https://en.wikipedia.org/wiki/Cohen%27s_kappa) -
  accuracy scaled by what is expected by random chance

Cohen's kappa will be 1 for perfect classification, 0 for no better
than random guessing and negative values as well, meaning the
prediction is somehow worse than random guessing.

# Using caret for training and evaluating models

In the machine learning notes, we will leverage an R package called
*caret* which provides a unified interface to many R implementations
of machine learning algorithms. *caret* stands for Classification And
REgression Training, meaning that the package can be used for
predicting categorical (classification) or continuous data
(regression). While there may be reasons to use the individual
packages themselves, *caret* gives us control of many functions in
separate R packages (the degree of control varies across package),
while wrapping them up in a consistent way so that users are less
likely to make mistakes, and code is easier to read.

**Importantly**: there are also a number of packages which are native
to languages other than R, such as the [Keras](https://keras.io/) deep
learning library in Python for construction of neural networks. While
there is limited support for constructing Keras models via *caret*, it
would make sense to instead use Keras in Python or using the R *keras*
package rather than the *caret* package for building certain
models. And as you go deeper into research in machine learning, and
development of new methods, you will likely switch to C or C++ for the
algorithm itself. In this last module of the course, we will not cover
developing new machine learning methods, but instead show how the
methods work (sometimes using R code to give a sense of the
calculations being performed) and show how many algorithms can be fit
within R using *caret*.

Within *caret* there are numerous functions for splitting data, and for
evaluating and comparing models. The author of the *caret* package,
Max Kuhn, has an
[online book for how to use caret](https://topepo.github.io/caret/index.html),
which is a useful reference for looking up, for example, 
[a list of all the machine learning algorithms available](https://topepo.github.io/caret/available-models.html).

Let's start with a very simple example of using *caret* to run a
linear model. Instead of using the `lm` function, we provide the
predictor variables (sometimes called features) `x` and the outcome
(or target) `y` to a function called `train` and specify
`method="lm"`. We can also pass additional arguments using the
`trainControl` function which is provided to the `trControl`
argument. `lm` is a kind of funny example, because it doesn't have any
*tuning parameters* whereas most of the other methods will have some
number of parameters that control behavior of the model, for example
limits or penalizations for model complexity. *caret* actually lists a
trivial parameter (`intercept=TRUE` or `FALSE`) for `lm`.

We will create some random Normally distributed data to demonstrate
the basic concept of overfitting. We can see that despite there being
no relationship between the outcome and the predictors (`y` consists
simply of 1,000 random Normal data points), we can obtain a strong
correlation between the predicted values. This is one of the main
concerns of machine learning methods: to find a way to combine
information in the predictors that will *generalize* to new data,
while not getting "tricked" into finding spurious patterns or
associations.

```{r include=FALSE}
knitr::opts_chunk$set(cache=TRUE)
```

```{r}
library(caret)
n <- 1000 # observations
p <- 500 # features
x <- matrix(rnorm(n*p),ncol=p,dimnames=list(seq_len(n),seq_len(p)))
y <- rnorm(n)
fit <- train(x, y, method="lm", trControl=trainControl(method="none"))
fit$results
plot(y, predict(fit))
cor(y, predict(fit))
```

Note the very high correlation, from having (1) many features and (2)
not so many more observations than features, relatively.

Above, `fit$results` provides no output because we specified in
`trainControl` that we will not use a resampling method to evaluate
performance. We can instead specify to use *cross-validation*, with 5
folds, and to save the predicted values. The default method for
resampling in *caret* is bootstrap sampling, where samples may appear
more than once in each split due to replacement, and therefore there
are out-of-boot samples that can be used for prediction. 

```{r}
trCtl <- trainControl(method="cv", number=5, savePredictions=TRUE)
fit <- train(x, y, method="lm", trControl=trCtl)
```

In case you have not yet seen cross-validation, the procedure is to
split the data into *k* roughly equally sized folds, and to then train
*k* different models. Each of the *k* models are different because
the input data to each model leaves out one of the folds. Each model
is therefore trained on roughly $n - \frac{n}{k}$ data points. The *k*
models can meanwhile be assessed by predicting the values of the
held-out data points. We use cross-validation in this example, but
note that other methods which involve repeated resampling help to
produce accuracy metrics with smaller standard errors.

There are many other methods for evaluation via resampling, which can
be found in the help page for `?trainControl` under `method`.

While we had a very high correlation when we fit a linear model to the
entire training set, here we see that the predictive $R^2$ on the held
out data in the cross-validation (CV) was nearly 0, indicating that
the high correlation above was due to overfitting.

```{r}
fit$results
head(fit$pred)
```

The `fit` object has a number of elements, which can be examined by
looking at `names(fit)`. The `fit$pred` element is a data.frame with a
column for the observed value of the outcome and a column for the
predicted value. Note that the observations are ordered by fold, not
by the original order in the dataset. 

What goes into the `fit$pred` table? If we had one or more tuning
parameters, there would be multiple copies of predicted values, e.g.
for each point in the tuning grid. We could pull out predictions in
that case by specifying the value(s) of the tuning parameter(s). Here,
as we did not have a tuning parameter, we only have a single copy of
each observation in the `fit$pred` table. Additionally, depending on
the resampling `method` specified in `trainControl`, we could have
multiple values per observation, per point in the tuning grid. For
`"boot"` and `"repeatedcv"` for example, an observation can appear
more than once in `fit$pred`.

By default, *caret* uses the squared-correlation definition of $R^2$
which we confirm by manually calculating the value (there would be no
reason to do this, except to convince oneself of what value is being
presented in the table above).

```{r}
getR2 <- function(pred) {
  r2s <- sapply(1:5, function(i) {
    idx <- pred[,"Resample"] == paste0("Fold",i)
    cor(pred[idx,2], pred[idx,1])^2
  })
  mean(r2s)
}
getR2(fit$pred)
```

We can also plot the predicted values over the observed values, and
color by the fold, showing that the linear model fit to the $n -
\frac{n}{k}$ data points had no predictive power for the held out data
points. 

```{r}
library(ggplot2)
ggplot(fit$pred, aes(obs, pred, color=Resample)) + geom_point()
```

# Feature selection using outcome must be in the loop

Many of the methods in `caret` have built in *feature selection*
properties, meaning that they pick out useful features and do not use
features which are not helpful for predicting the outcome. This is in
some sense wrapped up in the functions, although there are ways to
control and extract the features that are selected. Before we proceed,
I want to show one example of how things can go very wrong in machine
learning, if feature selection occurs outside of the cross-validation
or resampling loop that is used for parameter tuning and
evaluation. One can be mislead into thinking that the predictions are
much better than they actually are.

Below we assess the correlation of all of the features with the
outcome `y`. We then look at the 80% quantile of the absolute value of
the correlations and chose the 20% of features with correlation above
this value to form a new set of predictors `x.filt`. Because the
feature selection occurred outside of the cross-validation loop, we
can expect to find exaggerated performance, even though the
performance is assessed in a "held-out" set (the held-out set was not
held out for assessing the correlations).

```{r}
cors <- cor(x, y)[,1]
q <- quantile(abs(cors), .8)
x.filt <- x[,abs(cors) > q]
fit <- train(x.filt, y, method="lm", trControl=trCtl)
fit$results
```

Note that we get a small-to-moderate $R^2$ (not so high, but clearly
above 0), and the plot of predictions over the true, held-out data
points looks decent, although we know from construction that the
predictors have no relationship to the outcome, and therefore offer no
utility in prediction:

```{r}
ggplot(fit$pred, aes(obs, pred, color=Resample)) + geom_point()
```

To repeat, many of the methods we will use will perform their own
feature selection using the outcome variable, but it will occur
*within the loop*. In some cases, we will perform feature selection
before running `train`, however we will not use the outcome variable
in these cases; we can, for example, remove features that have 0
variance, or perform dimension reduction and retain the top *N*
principal components.

# Parameter tuning

A convenient feature of *caret* is that it helps us to tune parameters
within the `train` function. For demonstration purposes, we will load
the `iris` dataset and show how `train` performs tuning for a
classification problem, here classifying the label of two species
based on the four numeric features. We add an additional set of 100
non-informative predictors to demonstrate the benefit of
regularization using the elastic net implementation in the *glmnet*
package.

```{r}
summary(iris)
idx <- 51:150
x <- cbind(iris[idx,1:4], matrix(rnorm(100*100),ncol=100))
x <- scale(x)
y <- droplevels(iris$Species[idx])
table(y)
```

We build a grid of lambda (the regularization parameter) and alpha
(the mixing parameter, where alpha=1 corresponds to lasso or L1
regularization, and alpha=0 corresponds to ridge or L2
regularization). 

```{r}
tg <- expand.grid(lambda=c(.01,.05,.1,.25,.5,.75), alpha=c(.1,.25,.5,.75,1))
head(tg,8)
```

It is simple then to have `train` handle the running of models over a
grid of parameters. Note that if we had not set `tuneGrid`, a default
grid would be used. 

```{r}
fit <- train(x, y, method="glmnet", tuneGrid=tg)
ggplot(fit, metric="Kappa") + ylim(0,1)
```

We can ask for the best set of tuning parameters (simply the set the
gave the best performance as defined by `metric`, where a custom
performance metric could be provided).

```{r}
fit$bestTune
fit$results[rownames(fit$bestTune),]
```

Note that other schemes for choosing the best set of tuning parameters
exist other than choosing the maximum performance. These are discussed
in the 
[choosing the final model](https://topepo.github.io/caret/model-training-and-tuning.html#choosing-the-final-model)
section of the *caret* online book. For example, Breiman et al (1984)
suggests to pick the simplest model which is within one standard error
of the optimal performance (where standard errors are estimated using
resampling, as is available based on the resampling `method` we
choose). The reason that `train` by default chooses the best
performing model is likely explained by this quote from the book:

> The main issue with [rules such as within-one-standard-error] is
> related to ordering the models from simplest to complex. In some
> cases, this is easy (e.g. simple trees, partial least squares), but
> in cases such as this model, the ordering of models is subjective.

# Training vs test set evaluation

So far we have discussed using either cross-validation or the
out-of-boot samples for evaluation of model performance. This is
useful, and helps as we saw earlier to avoid thinking we have a much
higher predictive performance than we actually do. However, because we
look at many models, and then choose the best performing one, it is
standard to use a wholly separate *test set* in order to evaluate the
final model. The test set can be specified at the beginning, making
sure to balance the labels equally across the split. More
complex
[data splitting](https://topepo.github.io/caret/data-splitting.html) 
procedures are also available. Typically, between 20%-30% of the
dataset might be held-out for testing.

The full training and testing setup would then involve, creating a
training set index:

```{r}
train.idx <- createDataPartition(y, p=.75, list=FALSE)
# just demonstrating that the labels are balanced:
table(seq_len(100) %in% train.idx, y)
```

Splitting the data:

```{r}
train.x <- x[train.idx,]
train.y <- y[train.idx]
test.x <- x[-train.idx,]
test.y <- y[-train.idx]
```

Fitting the model and finding the best out-of-boot performing model: 

```{r}
fit <- train(train.x, train.y, method="glmnet", tuneGrid=tg)
fit$results[rownames(fit$bestTune),]
```

Assessing performance on the test set:

```{r}
pred <- predict(fit, newdata=test.x)
table(pred, test.y)
```

We can use a *caret* function to provide numerous metrics, including
accuracy and Cohen's kappa:

```{r}
confusionMatrix(data=pred, reference=test.y)
```

Or for less readout:

```{r}
postResample(pred=pred, obs=test.y)
```

# Bagging and boosting

*Bagging* and *boosting* are two important topics that come up across
a number of machine learning methods, and so I want to introduce them
during this introductory lecture note. *Bagging* is short for
*bootstrap aggregating*, wherein the potentially high variance of a
simple learning algorithm (or "learner") is mitigated by training it
multiple times across bootstrap samples of the dataset. In addition,
bagging may involve random sampling of the predictors as well, such
that the different models are less correlated than they would be if
they all used the same predictors. Bagging was proposed by Leo Breiman
in a
[technical report](https://www.stat.berkeley.edu/~breiman/bagging.pdf)
in 1994, which was subsequently published in 
[Machine Learning](https://doi.org/10.1007/BF00058655). The final
bagged model is an average over the simple models learned on the
bootstrap samples (either outputting the average prediction for
regression, or using a majority-rule vote for classification). Bagging
is very often used with *decision trees* which will be covered in more
detail in a subsequent lecture note. It is a natural question, why is
bagging not used with other simple models, like linear regression.  A
1998 paper by Skurichinaa and Duina,
[Bagging for linear classifiers](https://doi.org/10.1016/S0031-3203(97)00110-6),
looks into this question, showing that linear classifiers are in the
class of stable classifiers that don't benefit from bagging. Breiman
makes this point in his initial paper, that bagging is mostly useful
for unstable classifiers.

*Boosting* is somewhat similar to bagging, in that a final
meta-learner is constructed which should perform better than each
individual learner. However, instead of the simple learners being
different according to the bootstrap sample of the observations, in
boosting, a series of learners are constructed. The series is
constructed so that (after the first round), the weights for
mis-classified samples from the previous round are up-weighted. At a
given round, the previous round's classifier and a new simple learner
are added together, with a coefficient on the new simple learner
chosen to minimize the error in the training set. We will show in
practice how this procedure leads to improved predictions. 

## Contrived example of bagging

We can implement bagging in simple R code, and then show how it can be
performed with the *caret* library. We will create a very
unstable learner for our demonstration, to highlight the benefit of
bagging. As we said above, typically bagging is used in combination
with a decision tree as the learner. Suppose our data looks as
follows: 

```{r}
makeData <- function(n) {
  sd <- .5
  x1 <- c(rnorm(.5*n,-.5,sd),rnorm(.5*n,.5,sd))
  x2 <- c(rnorm(.5*n,-.5,sd),rnorm(.5*n,.5,sd))
  idx <- sample(n, .05*n)
  x1[idx] <- -5 * x1[idx]
  y <- factor(rep(c(1,2),n*c(.5,.5)))
  data.frame(x1,x2,y)
}
set.seed(1)
dat <- makeData(100)
y <- dat$y
x <- dat[c("x1","x2")]
ggplot(dat, aes(x1,x2,col=y)) + geom_point()
```

We can see that we have a group of observations that are outlying to
some degree. We will create a very unstable learner (just for
demonstration) which puts a 10x weight on the left- and right-most
observations, and then performs LDA.

```{r}
library(MASS)
learner <- function(x,y) {
  idx <- c(which.max(x[,1]), which.min(x[,1]))
  wt <- 10
  x <- rbind(x, do.call(rbind, lapply(seq_len(wt), function(i) x[idx,])))
  y <- c(y, rep(y[idx],wt))
  lda(x, y)
}
```

We can run it on our training data and plot the decision boundary. We
can see that the up-weighting of the extreme points leads to a poor
boundary for the majority of the points.

```{r}
l <- learner(x,y)
s <- seq(from=-5,to=5,length.out=200)
s.wide <- seq(from=-7,to=7,length.out=200)
grid <- expand.grid(x1=s.wide,x2=s)
grid.dat <- data.frame(y=predict(l, grid)$class, grid)
ggplot(dat, aes(x1,x2,col=y)) + geom_point() + 
  geom_raster(data=grid.dat, aes(x1,x2,fill=y),alpha=.2,show.legend=FALSE)
```

Now we write a simple function to predict the classes of the training
points: 

```{r}
learnerPred <- function(x,l) {
  predict(l, newdata=x)$class
}
table(pred=learnerPred(x,l), obs=y)
```

A bagging version of the learner involves bootstrap resampling of the
dataset. In some of the bootstrap samples, the most extreme points in
the dataset will not be included, and so some of the learners will
hopefully provide a better decision boundary for the majority of the
samples. Again, remember, our learner is contrived in a way to make it
very unstable, but you can imagine how individual data points might
lead an unstable learner to overfit the training data.

```{r}
B <- 400
reps <- lapply(seq_len(B), function(i) {
  boot <- sample(nrow(x), nrow(x), replace=TRUE)
  learner(x[boot,], y[boot])
})
```

Our "bagged" predictions simply involve running each of the *B* LDA
models and then using majority rules to vote for the final class. Note
that we do a bit better even on the training data.

```{r}
bagged <- sapply(seq_len(B), function(i) {
  learnerPred(x, reps[[i]])
})
bag.pred <- ifelse(rowSums(bagged == "2") > B/2, "2", "1")
table(pred=bag.pred, obs=y)
```

Now for some test data, we create 2,000 new observations. It's very
clear that the single learner is doing poorly here:

```{r}
dat.test <- makeData(2000)
s.wide <- seq(from=-10,to=10,length.out=200)
grid <- expand.grid(x1=s.wide,x2=s)
grid.dat <- data.frame(y=predict(l, grid)$class, grid)
ggplot(dat.test, aes(x1,x2,col=y)) + geom_point() +
  geom_raster(data=grid.dat, aes(x1,x2,fill=y),alpha=.2,show.legend=FALSE)
```

We can sum up the errors:

```{r}
y.test <- dat.test$y
x.test <- dat.test[,c("x1","x2")]
(tab <- table(pred=learnerPred(x.test,l), obs=y.test))
tab[1,2] + tab[2,1]
```

Now we can show that the bagging version of our unstable learner does
in fact do better: 

```{r}
bagged <- sapply(seq_len(B), function(i) {
  learnerPred(x.test, reps[[i]])
})
bag.pred <- ifelse(rowSums(bagged == "2") > B/2, "2", "1")
(tab <- table(pred=bag.pred, obs=y.test))
tab[1,2] + tab[2,1]
```

## Example of bagging trees

In our example above, we contrived a very unstable learner, which
up-weighted outliers, and then wrote out simple R code to show what
happens inside of a bagging method. However, we can easily access 
[bagging](https://topepo.github.io/caret/train-models-by-tag.html#bagging)
models via *caret*, including Random Forests, which we will cover in a
subsequent lecture note. This hides away the bootstrapping and
majority voting code, and there are a number of bagging methods, as
well as a general purpose `bag` interface within *caret*.

Below we use a bagging version of decision trees, which gets close to a
Random Forest, but doesn't involve sub-sampling of the features (here
we only have two features anyway). If you think of individual decision
trees as dividing the predictor space into blocks where a certain
class will be predicted (for classification, or a level value for
regression), then by averaging over many trees, complex decision
boundaries can be learned. Here we show how a bagging version of
decision trees can be applied:

```{r}
fit <- train(x, y, method="treebag")
fit$results
pred <- predict(fit, x.test)
tab <- table(pred, obs=y.test)
tab[1,2] + tab[2,1]
```

Note that our bagging method *treebag* outperforms simple LDA:

```{r}
fit <- train(x, y, method="lda")
fit$results
pred <- predict(fit, x.test)
tab <- table(pred, obs=y.test)
tab[1,2] + tab[2,1]
```

## Example of boosting

As we have limited time in the course, we do not cover boosting in
depth, but only provide a demonstration of its use. As described
above, boosting involves iteratively re-fitting a learner, typically a
tree, wherein the previous learners in the series are summed up and a
new learner is added with additional focus on observations in the
training set that we mis-classified in the previous round. As with
bagging, this can lead to complex behavior and complex decision
boundaries from a series of simple learners.

A very powerful boosting method is GBM, for 
[Gradient Boosting Machine](https://en.wikipedia.org/wiki/Gradient_boosting) 
(link to Wikipedia).
It can be easily called within *caret* by specifying
`method="gbm"`. Another popular boosting method is 
[AdaBoost](https://en.wikipedia.org/wiki/AdaBoost) 
(link to Wikipedia),
which can be called with *caret* via `method="adaboost"`.

Here we apply GBM to our simple dataset for demonstration of how 
the learned surface may look. Typically we would use GBM on more
complex datasets with more than two dimensions, but also notice that
it outperforms the other methods on the test data:

```{r}
fit <- train(x, y, method="gbm", verbose=FALSE)
fit$results
pred <- predict(fit, x.test)
tab <- table(pred, obs=y.test)
tab[1,2] + tab[2,1]
```

We can plot, with the final model, the training error over the rounds
of boosting. At each round, the method is attempting to improve on
certain mis-classified samples, while still retaining good
classification on the "easy" samples:

```{r}
plot(fit$finalModel$train.error, type="o",
     xlab="round", ylab="train error")
```

Before we plot the learned surface, let's recall what the training
data looked like:

```{r}
ggplot(dat, aes(x1,x2,col=y)) + geom_point()
```

The `plot` function, called on the final model, can be used to make
low-dimensional plots showing how the variables relate to the class
predictions. From `?plot.gbm`:

> `plot.gbm` produces low dimensional projections of the
> `gbm.object` by integrating out the variables not included in the
> `i.var` argument. The function selects a grid of points and uses
> the weighted tree traversal method described in Friedman (2001) to
> do the integration.

Here we plot the surface for `x1`, for `x2` and in the space of both
variables. Note that, by adding over many simple learners, a complex
shape can be created. Still, these boosted methods are quite good at
not overfitting. GBM is one of the best methods for out-of-the-box
performance, that is not requiring much "feature engineering":

```{r}
plot(fit$finalModel, i.var=1)
plot(fit$finalModel, i.var=2)
plot(fit$finalModel, i.var=1:2)
```

# Complexity of a classifier

The final section of this lecture note will cover a formal definition
of complexity that is used in machine learning, and a useful theorem
giving a probabilistic upper bound on how far the test error will be
from the training error of a classifier.

First, we will define the concept of an algorithm *f* with parameters
$\theta$ *shattering* a set of points. To shatter a set of points
means that, for any labeling of the points, there exists a $\theta$
such that *f* classifies all the points accurately. For the following
examples, suppose we are working on binary classification, and for
simplicity of notation, we will have the two class labels as
$\{-1,+1\}$.

Suppose we consider algorithms *f* which are of the form:

$$ \textrm{sign}(a x_1 + b x_2 + c) $$

with $a,b > 0$ (and a 0 will be given a positive label). We can
visualize a particular instance of *a*, *b*, and *c*:

```{r echo=FALSE, fig.width=6}
a <- 3; b <- 1; c <- -2
f <- function(x) factor(sign(a*x[,1]+b*x[,2]+c+1e-6))
s <- seq(from=-4,to=4,length.out=200)
grid <- expand.grid(x1=s,x2=s)
grid$y <- f(grid)
ggplot(grid, aes(x1,x2,fill=y)) + geom_raster(alpha=.2) +
  geom_hline(yintercept=0) + geom_vline(xintercept=0)
```

Now let's lay down a single point and ask if our algorithm *f* as defined
above can shatter it. If we have a single positive point, we can
simply move the line so that it is below the point:

```{r echo=FALSE, fig.width=6}
dat <- data.frame(x1=-1,x2=-1,y=factor(1,levels=c(-1,1)))
a <- 1; b <- 2; c <- 4
grid$y <- f(grid)
ggplot(dat, aes(x1,x2,color=y)) +
  geom_point(shape=4,size=5,stroke=2, show.legend=FALSE) + 
  geom_raster(data=grid, aes(x1,x2,fill=y), alpha=.2) +
  geom_hline(yintercept=0) + geom_vline(xintercept=0)
```

Likewise, with a negative point, we can move the line above it:

```{r echo=FALSE, fig.width=6}
dat <- data.frame(x1=-1,x2=-1,y=factor(-1,levels=c(-1,1)))
a <- 1; b <- 2; c <- 1
grid$y <- f(grid)
ggplot(dat, aes(x1,x2,color=y)) +
  geom_point(shape=4,size=5,stroke=2, show.legend=FALSE) + 
  geom_raster(data=grid, aes(x1,x2,fill=y), alpha=.2) +
  geom_hline(yintercept=0) + geom_vline(xintercept=0)
```

Intuitively, *f* as defined above can shatter a single point.

Now, suppose we have two points, with one in the upper right and one
in the bottom left quadrant. We can correctly classify if the point
toward the upper right quadrant is positive:

```{r echo=FALSE, fig.width=6}
dat <- data.frame(x1=c(-1,1),x2=c(-1,1),y=factor(c(-1,1)))
a <- 1; b <- 1; c <- 0
grid$y <- f(grid)
ggplot(dat, aes(x1,x2,color=y)) +
  geom_point(shape=4,size=5,stroke=2, show.legend=FALSE) + 
  geom_raster(data=grid, aes(x1,x2,fill=y), alpha=.2) +
  geom_hline(yintercept=0) + geom_vline(xintercept=0)
```

...however there is no $\theta$ that will be able to correctly
classify both when the labels are reversed:

```{r echo=FALSE, fig.width=6}
dat <- data.frame(x1=c(-1,1),x2=c(-1,1),y=factor(c(1,-1)))
a <- 10; b <- 1; c <- 20
grid$y <- f(grid)
ggplot(dat, aes(x1,x2,color=y)) +
  geom_raster(data=grid, aes(x1,x2,fill=y), alpha=.2) +
  geom_point(shape=4,size=5,stroke=2, show.legend=FALSE) + 
  geom_hline(yintercept=0) + geom_vline(xintercept=0)
```

Let's now consider another set of two points. If the upper left
quadrant is negative, we can tilt the slope very steeply:

```{r echo=FALSE, fig.width=6}
dat <- data.frame(x1=c(-1,1),x2=c(1,-1),y=factor(c(-1,1)))
a <- 5; b <- 1; c <- 0
grid$y <- f(grid)
ggplot(dat, aes(x1,x2,color=y)) +
  geom_raster(data=grid, aes(x1,x2,fill=y), alpha=.2) +
  geom_point(shape=4,size=5,stroke=2, show.legend=FALSE) + 
  geom_hline(yintercept=0) + geom_vline(xintercept=0)
```

And if the point in the upper left quadrant is positive, we can tilt
the slope very shallowly:

```{r echo=FALSE, fig.width=6}
dat <- data.frame(x1=c(-1,1),x2=c(1,-1),y=factor(c(1,-1)))
a <- 1; b <- 5; c <- 0
grid$y <- f(grid)
ggplot(dat, aes(x1,x2,color=y)) +
  geom_raster(data=grid, aes(x1,x2,fill=y), alpha=.2) +
  geom_point(shape=4,size=5,stroke=2, show.legend=FALSE) + 
  geom_hline(yintercept=0) + geom_vline(xintercept=0)
```

So while *f* as defined above could not shatter the first set of
points (the ones on the NE-SW axis), *f* could shatter the second set
of points (the ones on the NW-SE axis).

If we lift the positive-valued restriction on *a,b*, you can work out
with examples how a real-valued version of *f* can shatter a set of
three points as long as they do not lie on a straight line.

Finally, the definition: The *VC dimension* of a classification
algorithm is the size of the largest set of points that the algorithm
can possibly *shatter*. There only has to be one set of such points,
and so the positive-valued version of *f* has VC dimension 2 (no set
of 3 points can be shattered) while the real-valued version of *f* has
VC dimension 3 (no set of 4 points can be shattered). As typical,
finding a set of points which an algorithm can shatter is the easy
part, while proving that no $\theta$ exists such that an algorithm can
shatter so many points is more difficult. *VC* stands for
Vapnik–Chervonenkis, who outlined the definition in a 1971 paper,
"*On the Uniform Convergence of the Frequencies of Occurrence of
Events to Their Probabilities*". 

Our main interest is in knowing the degree to which our ability to 
predict the labels of the training set will generalize to our ability
to predict new data that arises from the same distribution. Here, when
we talk about training set error, we do not refer to the
cross-validated or out-of-bag error, but the error on the same
observations that are used for finding, e.g. $\theta$ for *f*
above.

My preferred walk-through of the VC dimension, and its relation to the
bound on generalization error is 
[Andrew Ng's notes on Learning Theory](http://cs229.stanford.edu/notes/cs229-notes4.pdf). 
I would recommend to walk through these. In particular, these notes
show how the estimated error in the training set must be close to the
generalization error (error on new data). For a finite set of fitted
models (where a "fitted model" is a function that maps from the
predictor space to the space of labels of the outcome), the
generalization error can be bound to within $\gamma$ of the training
error, for all of the fitted models, with probability:

$$ P \ge 1 - 2k\exp(-2 \gamma^2 n)$$

where *k* is the number of models and *n* is the number of
samples. This bound uses the union bound and the Hoeffding inequality
in its derivation. Then using this probability bound, one can also
bound the generalization error for the *best* fitted model (in terms
of minimizing error on the training samples) among the *k*
options. 

However, we rarely work with finite sets of models, but instead work
with infinite sets of models, such as the *f* we were describing
above. Here, the VC dimension (if it is finite) becomes useful as way
of describing the complexity of the set. For a classification
algorithm with VC dimension *d*, with probability at least $1 -
\delta$ the absolute difference between generalization error and the
training error will be:

$$ \le O\left( \sqrt{\frac{d}{n} \log \frac{n}{d} + \frac{1}{n} \log \frac{1}{\delta}} \right) $$

Thus for classification algorithms with finite VC dimension, large
sample size will give us probabilistic guarantee on only small
differences between training and generalization error. Again for a
closer walk through of these bounds, I recommend consulting 
[Andrew Ng's notes on Learning Theory](http://cs229.stanford.edu/notes/cs229-notes4.pdf).
