---
title: "Neural networks"
author: "Michael Love"
date: 12/15/2018
output: html_document
---

This lecture note will give a brief introduction to (artificial)
neural networks (where we may include the word "artificial" to
contrast with biological neural networks). We begin by reviewing
logistic regression and the gradient descent/acsent method, and
connect this to the 
[perceptron](https://en.wikipedia.org/wiki/Perceptron).
We then describe how layers of perceptrons can lead to complex
prediction algorithms. Finally, we will construct various neural
networks for classifying the MNIST handwritten digits dataset. The
field of neural networks and deep learning is vast and quickly
evolving. For those interested in exploring the topic, I recommend
searching for current methods implementing deep learning for a
specific problem of interest.

This lecture note borrows from Andrew Ng's course notes for
[supervised learning](http://cs229.stanford.edu/summer2020/cs229-notes1.pdf), 
Ng and Katanforoosh's notes on 
[deep learning](http://cs229.stanford.edu/summer2020/cs229-notes-deep_learning.pdf),
and the (offline) machine learning course notes
of [Guenther Walther](http://statweb.stanford.edu/~gwalther/).

```{r include=FALSE}
knitr::opts_chunk$set(cache=TRUE)
```

# Gradient descent for LS and logistic regression

Recall that if we want to minimize least squares in a simple linear
model,

$$ \hat{y} = w_1 x_1 + w_2 x_2 + \dots w_p x_p + b, $$

we can use a gradient descent update rule for each predictor *j*,
either a *batch* update: 

$$ w_j^{t+1} = w_j^t + \lambda \textstyle \sum_{i=1}^N (y_i - \hat{y}_i) x_{ij}, $$ 

where we sum over all observations *i*, or *stochastic* gradient
descent where we update the $w_j$ by looking at one *i* at a time:

$$ w_j^{t+1} = w_j^t + \lambda (y_i - \hat{y}_i) x_{ij} $$ 

This is the gradient descent method, where we move in the direction
that minimizes the sum of squared error for each predictor, and where
the parameter $\lambda$ controls the rate of descent.

If we instead use the logistic function for a binary target,

$$ \hat{y} = f(w_1 x_1 + w_2 x_2 + \dots w_p x_p + b), $$

with a logistic function, *f*:

$$ f(z) = \frac{1}{1 + \exp(-z)}, $$

we have the same update rule for gradient descent of the negative log
likelihood (here the stochastic version),

$$ w_j^{t+1} = w_j^t + \lambda (y_i - \hat{y}_i) x_{ij}, $$ 

where $y_i \in \{ 0,1 \}$ and $\hat{y} \in [0,1]$. We will see how
we can build up from these relatively simple algorithms using similar
update rules to learn more complex relationships between predictors
and target. 

# Perceptron

A limiting case of the logistic function is a function that only
outputs values of 0 or 1 around a critical value of $z = 0$:

```{r echo=FALSE, fig.width=5}
plot(function(x) 1/(1 + exp(-x)), xlim=c(-5,5), xlab="z", ylab="f(z)")
plot(function(x) 1/(1 + exp(-2*x)), xlim=c(-5,5), add=TRUE)
plot(function(x) 1/(1 + exp(-10*x)), xlim=c(-5,5), add=TRUE)
plot(function(x) ifelse(x >= 0, 1, 0), xlim=c(-5,5), col="dodgerblue", lwd=3, add=TRUE)
```

If we use this step function instead of the logistic, the
classification algorithm is called the 
[perceptron](https://en.wikipedia.org/wiki/Perceptron).
The perceptron was invented in 1957 at the Cornell Aeronautical
Laboratory. From Wikipedia:

> The perceptron was intended to be a machine, rather than a program,
> and while its first implementation was in software for the IBM 704,
> it was subsequently implemented in custom-built hardware as the
> "Mark 1 perceptron". This machine was designed for image
> recognition: it had an array of 400 photocells, randomly connected
> to the "neurons". Weights were encoded in potentiometers, and weight
> updates during learning were performed by electric motors. 

For the perceptron, again we can use the update rule:

$$ w_j^{t+1} = w_j^t + \lambda (y_i - \hat{y}_i) x_{ij} $$

The [course notes from Michael Collins](http://www.cs.columbia.edu/~mcollins/courses/6998-2012/lectures/lec1.3.pdf) 
on the perceptron contain a useful proof that if there exists a set of
parameters such that all training samples can be classified correctly,
then the above algorithm will converge.

Although the update rules look very similar, the following note from
Andrew Ng's notes on supervised learning is apt:

> Note however that even though the perceptron may
> be cosmetically similar to the other algorithms we talked about, it
> is actually a very different type of algorithm than logistic regression and
> least squares linear regression; in particular, it is difficult to
> endow the perceptron’s predictions with meaningful probabilistic
> interpretations, or derive the perceptron as a maximum likelihood
> estimation algorithm.

Here we construct a simple case and show how the perceptron using the
above update rule moves to reduce training errors.

```{r}
set.seed(1)
n <- 100
x1 <- c(rnorm(n),rnorm(n,4))
x2 <- c(rnorm(n),rnorm(n,4))
x1 <- scale(x1)
x2 <- scale(x2)
y <- rep(0:1,each=n)
dat <- data.frame(y=factor(y),x1,x2)
```

Here we encode the update rule, and run five iterations:

```{r}
# pick initial values for w and b
w <- c(-4,1)
b <- 3

# create matrices to track w and b for 6 iterations/updates
w.mat <- matrix(nrow=6,ncol=2)
b.vec <- numeric(6)

# choose learning rate
lambda <- 0.01

# run the for loop
for (i in 1:6) {
  # save current w and b
  w.mat[i,] <- w
  b.vec[i] <- b
  
  # get yhat given current w and b via perceptron
  y.hat <- ifelse(cbind(x1,x2) %*% w + b >= 0, 1, 0)
  
  # update each w (two)
  for (j in 1:2) {
    w[j] <- w[j] + lambda * sum((y.hat - y) * x1)
  }
  
  # update b
  b <- b + lambda * sum(y.hat - y)
}

# save final w and b
w.mat[i,] <- w
b.vec[i] <- b

# create data frame for plotting
dat2 <- data.frame(b=b.vec, w1=w.mat[,1], w2=w.mat[,2], iter=1:6)
```

We can plot the decision boundary at each iteration:

```{r fig.width=5.5}
library(ggplot2)
ggplot(dat, aes(x1,x2,shape=y)) + geom_point() +
  scale_shape_discrete(solid=FALSE) +
  geom_abline(data=dat2, aes(intercept=-b/w2, slope=-w1/w2, color=iter), size=1) +
  scale_color_continuous(low="purple",high="orange")
```

It is clear that a classifier of the form of the perceptron will only 
generate linear decision boundaries. However, we can add multiple
layers of perceptrons to form a more complex classifier. 

First, let's visualize the kind of contours we obtain when we go straight from
inputs to the target with a single perceptron, assuming $w = (2, -1)$ and $b = 1$:

```{r fig.width=5.5}
s <- seq(from=-2,to=2,length=400)
# grid spanning  cominbations of x1 and x2 given s
grid <- expand.grid(x1=s,x2=s)
# predict y's across  grid region
grid$y <- with(grid, x1 * 2 + x2 * -1 + 1)
ggplot(grid, aes(x1,x2,z=y,fill=y)) +
  geom_raster(alpha=.5) +
  scale_fill_continuous(low="blue",high="green") +
  geom_contour()
```

Now, instead of a single set of weights from the input variables
directly predicting *y*, we can add a hidden layer in between the
inputs and the output. Suppose we add four nodes in between, giving us
eight weights from the inputs to the hidden nodes (plus four
bias/intercept terms), and four weights from the hidden nodes to the
target (plus one bias term). What kind of shapes might be recoverable
with such a network?

```{r fig.width=5.5}
s <- seq(from=-2,to=2,length=400)
# creating grid again
grid <- expand.grid(x1=s,x2=s)
# this is same as the previous example, except now we use the perceptron
# there are 4 nodes in the hidden layer, each with separate w's, between x  and the predicted y
# node 1, w = (1,  1), b = 1
grid$h1 <- with(grid, ifelse(1*x1 + 1*x2 + 1 >= 0, 1, 0))
# node 2, w = (1,  -1), b = 1
grid$h2 <- with(grid, ifelse(1*x1 - 1*x2 + 1 >= 0, 1, 0))
# node 3, w = (-1,  1), b = 1
grid$h3 <- with(grid, ifelse(-1*x1 + 1*x2 + 1 >= 0, 1, 0))
# node 4, w = (-1,  -1), b = 1
grid$h4 <- with(grid, ifelse(-1*x1 - 1*x2 + 1 >= 0, 1, 0))
# final linear output node, taking input the 4 nodes from the hidden layer to predict y
# w = (1, 1, 1, 1) here, taking the output from each hidden layer node and b = -3.5
grid$y <- with(grid, 1*h1 + 1*h2 + 1*h3 + 1*h4 - 3.5)
# plot
ggplot(grid, aes(x1,x2,z=y,fill=y)) +
  geom_raster(alpha=.5) +
  scale_fill_continuous(low="blue",high="green") +
  geom_contour()
```

The hidden layer adds a round of intermediate transformations of the input variables.  The outputs from these nodes are then passed on to a final output layer for a final round of transformation.  By applying a series of simple linear transformations, one can achieve very complex decision boundaries.  

In practice, the values for the weights and biases in each node are unknown, and are therefore learned with gradient descent.  More complex decision boundaries can be learned with adding additional nodes and hidden layers (the latter is where the "deep" in deep learning)

There is a much more elegant visualization from *Tensorflow* of how
multiple layers of nodes/neurons can be used to learn complex
decision boundaries for various datasets (here with 2D input for ease
of visualization):

* [Tensorflow Playground](http://playground.tensorflow.org)

Having gotten a sense for how multiple layers of nodes can be used to
learn complex patterns, a natural question is how we will be able to
fit the weights for these multi-layer perceptrons (MLP). 

What if we attempt our approach from earlier: define a loss function, and then take the partial derivative of this with respect to each weight, and move some direction down the gradient at
each iteration. There are numerous problems with the gradient descent approach here: 
1. Certain *activation functions* such as the step function we had above and the Rectified linear unit (ReLU) function, as well as the addition of hidden layers, make taking the derivative difficult, and 
2. Here will be many local minima where the descent method gets stuck.

Instead, a method called *backpropagation* can be used. This method
starts with the layer closest to the target *y* and, one weight at a
time, takes the partial derivative with respect to this weight. Then
the weight can be adjusted using the old weight and multiplying the
gradient by the learning rate $\lambda$. The weights for the nodes in
the layer second closest to the target are then updated, which can be
computed with a trick that uses the chain rule to reformulate a product
of terms that can be calculated. The name for the method stands for
"backward propagation of error" because the derivative of the error
from the layers closer to the target are propagated backward through
the network recursively. We do not derive or demonstrate
backpropagation in this note (it is derived in the notes from Ng and
Katanforoosh on
[deep learning](http://cs229.stanford.edu/summer2020/cs229-notes-deep_learning.pdf) and
[backpropagation](http://cs229.stanford.edu/notes-spring2019/backprop.pdf),
but instead will describe a number of options when building neural networks, before demonstrating the use of one in classifying
handwritten digits:

# Activation functions

The term *activation functions* describe the functions $f(z)$ that are
applied to the weighted combination of the incoming nodes. Above we
discussed a *logistic*, or *sigmoid*, activation function, as well as a
step function $f(x) = 1_{\{x \ge 0\}}$. There are numerous choices,
which can be browsed at the following 
[Wikipedia page](https://en.wikipedia.org/wiki/Activation_function). 
Properties of an activation function to consider are whether they have
a continuous derivative, the steepness of the curve, whether it
asymptotes for large absolute values of the input, and the range. We
plot some of the more popular activation functions, including ones
which we will use in the example at the end of this note.

```{r echo=FALSE, fig.width=6}
par(mfrow=c(2,2),mar=c(3,3,2,1))
myplot <- function(f, main) {
  plot(f, main=main, xlim=c(-3,3), ylim=c(-2,2), xlab="", ylab="", type="n")
  abline(h=0)
  plot(f, col="dodgerblue", lwd=3, add=TRUE, xlim=c(-3,3))
}
myplot(function(x) 1/(1 + exp(-x)), "logistic")
myplot(function(x) tanh(x), "tanh")
myplot(function(x) ifelse(x >= 0, x, 0), "ReLU")
myplot(function(x) ifelse(x > 0, x, (exp(x) - 1)), "ELU")
```

In addition, the *softmax* activation function is useful for mapping
from a vector defined in $\mathcal{R}^n$ to a vector on the
(n-1)-simplex defined such that each element is restricted between 0
and 1 and the elements sum to 1:

$$ f_i(\vec{z}) = \frac{\exp(z_i)}{\sum_{j=1}^J \exp(z_j)} $$

This activation function will be useful when we are trying to predict
multiple classes, such as the handwritten digits data (0-9) we will
consider below.

# Regularization / weight decay

As with many classification or regression models with many parameters
to learn, the models can become unstable and potentially have
difficulties with overfitting. We therefore may benefit from
considering *regularization*, that is, the penalization of large
weights. As with penalized regression, we can consider an L1 or L2
regularization on the weights, as well as the tuning between the loss
function and the penalty term. Another regularization implementation is
to have the weights slowly decay during training iterations, by simply
multiplying the weights by a value less than 1 at each iteration. It
turns out that weight decay is 
[equivalent to L2 regularization](https://arxiv.org/abs/1711.05101)
for certain training methods (stochastic gradient descent) but not for
all algorithms, which should be kept in mind.

# Convolutional layers

A final topic which will be important for our application below is the
concept of convolutional layers of nodes in the beginning of a neural
network. These are important component of networks that are designed
to take images or 3D volumes as input, where the features $x_j$ have
some meaningful spatial relationship that can be captured by features
learned in the hidden layers. A good extended reference for
convolutional neural networks is the
[CNN lecture note by Andrej Karpathy](http://cs231n.github.io/convolutional-networks/).

One motivation of convolutional layers is that we may want hidden
layers of the network to be able to detect common patterns across the
image, regardless of where they occur positionally in the image. We
can convolve a *filter* across the pixels of the image (or voxels of a
volume) and take the sum as the output. As explained best by the
lecture note linked above:

> Intuitively, the network will learn filters that activate when they
> see some type of visual feature such as an edge of some orientation or
> a blotch of some color on the first layer, or eventually entire
> honeycomb or wheel-like patterns on higher layers of the network.

In addition *pooling* layers are typically added as well which
additionally reduce the input to reduce the number of parameters to
learn. One way to reduce the image is to take the maximum of every 2x2
sub-matrix and pass this through to the next layer. 

# MNIST handwritten digits

We will now explore a dataset where neural networks provide a clear
advantage in prediction. The 
[MNIST handwritten digits dataset](https://en.wikipedia.org/wiki/MNIST_database)
is a famous one for machine learning algorithms. MNIST stands for
"Modified National Institute of Standards and Technology". From
Wikipedia:

> [MNIST] was created by "re-mixing" the samples from NIST's original
> datasets. The creators felt that since NIST's training dataset was
> taken from American Census Bureau employees, while the testing
> dataset was taken from American high school students, it was not
> well-suited for machine learning experiments. Furthermore, the
> black and white images from NIST were normalized to fit into a 28x28
> pixel bounding box and anti-aliased, which introduced grayscale
> levels.
> 
> The MNIST database contains 60,000 training images and 10,000
> testing images. Half of the training set and half of the test set
> were taken from NIST's training dataset, while the other half of the
> training set and the other half of the test set were taken from
> NIST's testing dataset. There have been a number of scientific
> papers on attempts to achieve the lowest error rate; one paper,
> using a hierarchical system of convolutional neural networks,
> manages to get an error rate on the MNIST database of 0.23%.

We will begin by reading in the data and visualizing some of the
digits. We use a CSV file created by Joseph Redmon, and some code from
David Robinson's blog post, both linked below:

* <https://pjreddie.com/media/files/mnist_train.csv>
* <http://varianceexplained.org/r/digit-eda/>

The following R code, from David Robinson's blog post, re-organizes
the raw data so that we can visualize the digits in a 2D matrix. The
predictors are the 28 x 28 = 784 pixels in the image, in grayscale,
with values from 0 to 255. The target is the digit itself, which takes
values from 0-9. In the CSV file, the first column is the target, and
the remaining 784 columns are the grayscale values of the pixels. The
predictors are ordered row-by-row, so the first 28 columns after the
target column make up the first row of the image. The modulo
arithmetic below gives us x- and y-coordinates for plotting the
pixels. 

```{r}
library(readr)
library(tidyr)
library(dplyr)
memory.limit(size=56000)
gc()
# read in data
mnist_raw <- read_csv("mnist_train.csv", col_names=FALSE)

# check dimensions and a few observations
dim(mnist_raw)
mnist_raw[1:5,1:5]

# reshape into table with columns as the following
  # label
  # obs ID (instance)
  # pixel number
  # value of pixel
  # x coordinate of pixel in image
  # y coordinate of pixel in image

pixels_gathered <- mnist_raw %>%
  head(10000) %>% # take 1st 10000 obs
  rename(label = X1) %>%
  mutate(instance = row_number()) %>%
  gather(pixel, value, -label, -instance) %>%
  tidyr::extract(pixel, "pixel", "(\\d+)", convert = TRUE) %>%
  mutate(pixel = pixel - 2,
         x = pixel %% 28,
         y = 28 - pixel %/% 28)
gc()
```

```{r}
library(ggplot2)
sub <- pixels_gathered %>% filter(instance <= 12)
ggplot(sub, aes(x, y, fill = value)) + geom_tile() +
  facet_wrap(~ instance + label, labeller=label_both) +
  scale_fill_gradient(low="white", high="black")
```

We will now forget for a while that the predictors make up pixels in a
2D matrix (until the end when we will use convolutional layers). We
will for the meantime just consider the predictors as exchangeable
and not spatially oriented. 

Before we try to predict all 10 digits, let's try just trying to
predict the digits 3, 5 and 8. We will see that even these three
digits are not trivial to classify: 

```{r}
# drop label and normalize values to between 0 and 1
dat <- as.data.frame(mnist_raw[1:10000,-1])/255

# convert label to factor 
y <- factor(mnist_raw$X1[1:10000])

# subset only to images with labels 3, 5, or 8
idx <- y %in% c("3","5","8")
dat <- dat[idx,]
y <- droplevels(y[idx])

# pca
pc <- prcomp(dat)
```

We can reduce the dimensionality of the data substantially by taking
just the top PCs:

```{r}
plot(pc$sdev[1:20]^2/sum(pc$sdev^2))
```

Just taking the top 7 PCs, we can see there is some separation along,
e.g. PC1, or looking at PC1 vs PC2:

```{r}
x <- pc$x[,1:7]
boxplot(x[,1] ~ y)
ggplot(data.frame(x,y), aes(PC1, PC2, col=y)) + geom_point()
table(y)
```

A very simple method for classification is *k-nearest neighbors*,
where we calculate the Euclidean distance of each sample in the test
or out-of-bag set to those samples in the training set. For classification problems, a majority
vote of the closest *k* training samples becomes the prediction.
The difficulties for *kNN* are that distances in high-dimensional space
become problematic, due to what is called the 
[curse of dimensionality](https://en.wikipedia.org/wiki/Curse_of_dimensionality#Distance_functions).

Briefly, it becomes difficult to get a good sense of neighbors in high
dimensional space. Another problem is that the algorithm is not
sparse: we have to save the entire training dataset in order to
classify new points. However, it is very straightforward to implement,
so it makes sense as a first try (or a lower bound in terms of
accuracy which we will try to improve upon). 

```{r}
library(caret)
tg <- data.frame(k=c(5,9,13,17,21,25))
trCtl <- trainControl(savePredictions=TRUE)
fit <- train(x, y, method="knn", tuneGrid=tg, trControl=trCtl)
```

We can plot the kappa statistics for various settings of *k*:

```{r}
ggplot(fit, metric="Kappa")
```

Also the confusion matrix, to see which digits are being misclassified:

```{r}
fit$results
tab <- table(obs=fit$pred$obs, pred=fit$pred$pred)
prop <- tab/rowSums(tab)
round(prop, 3) * 100 # percent
```

Rather than looking at a single kappa statistics, we will write this
small function to plot 2x the standard deviation of the statistic,
over bootstrap replications:

```{r}
plotWithSD <- function(fit, param, a=2) {
  min <- with(fit$results, min(Kappa - (a+2)*KappaSD))
  max <- with(fit$results, max(Kappa + (a+2)*KappaSD))
  fit$results$ymax <- with(fit$results, Kappa + a*KappaSD)
  fit$results$ymin <- with(fit$results, Kappa - a*KappaSD)
  ggplot(fit$results, aes_string(param, "Kappa", ymax="ymax", ymin="ymin")) +
    geom_ribbon(fill="black", alpha=.1) +
    geom_point(color="blue") + geom_line(color="blue") +
    ylim(min,max)
}
```

We see that the kappa statistic plateaus around 0.85 even though we
are only trying to classify the digits 3, 5, and 8.

```{r}
plotWithSD(fit, "k")
```

Now let's return to all the digits and see how we do with our kNN
approach: 

```{r}
dat <- as.data.frame(mnist_raw[1:10000,-1])/255
y <- factor(mnist_raw$X1[1:10000])
pc <- prcomp(dat)
```

Again, looking at PC1 and PC1 vs PC2, we see some separation of the
digits already. We will work with the top 10 PCs (chosen arbitrarily,
mostly for computational speed of training the methods below).

```{r}
x <- pc$x[,1:10]
boxplot(x[,1] ~ y)
ggplot(data.frame(x,y), aes(PC1, PC2, col=y)) + geom_point()
```

We can again try kNN and we see that it's doing fairly well across
all 10 digits, with a kappa statistic that plateaus at around 0.885.

```{r}
tg <- data.frame(k=c(5,9,13,17,21,25))
trCtl <- trainControl(savePredictions=TRUE)
fit.knn <- train(x, y, method="knn", tuneGrid=tg, trControl=trCtl)
fit.knn$results # 88% Kappa
plotWithSD(fit.knn, "k")
```

It's useful to look at the confusion matrix:

```{r}
tab <- table(obs=fit.knn$pred$obs, pred=fit.knn$pred$pred)
prop <- tab/rowSums(tab)
round(prop,3)*100 # percent
diag(prop) <- NA
```

We can plot this, after removing the diagonal elements, and highlight
the top three mistakes that are made by our kNN approach. We see
that 4 and 9 are often misclassified as each other, as well as 3
and 8. 

```{r fig.width=5, fig.height=5.5}
image(prop, xaxt="n", yaxt="n",
      xlab="obs", ylab="pred",
      col=colorRampPalette(c("white","blue"))(50))
for (i in 1:2) axis(i, 0:9/9, 0:9)
abline(0,1,col="red")
mxs <- head(sort(prop,decreasing=TRUE),3)
lbl <- 100*round(mxs,3)
arr.idx <- sapply(mxs, function(i) which(prop == i, arr.ind=TRUE)) - 1
text(arr.idx[1,]/9, arr.idx[2,]/9, labels=lbl, cex=.9, col="white")
```

We can also try a linear SVM. A radial basis function SVM does
out-perform this but takes about 15 times longer to fit (we observed
kappa statistic up to 0.94). The linear SVM doesn't do much better
than the kNN but is more efficient; it doesn't store the whole
dataset, only the support vectors (although here we have 2,000+
support vectors).

```{r}
x <- pc$x[,1:20]
fit <- train(x, y, method="svmLinear", trControl=trCtl)
fit$results
fit$finalModel@nSV
```

# Multi-layer perceptron

There are a few neural networks that can be accessed via *caret* but
we will show that these underperform compared to a fully specified
neural network with convolutional layers. Here, we take the top 60 PCs
and train a multi-layer perceptron (MLP) with a single hidden layer
with 60 nodes.

```{r}
x <- pc$x[,1:60]
tg <- data.frame(size=60)
trCtl <- trainControl(method="cv", number=5, savePredictions=TRUE, verboseIter=TRUE)
fit <- train(x, y, method="mlp", trControl=trCtl, tuneGrid=tg, maxit=50)
```

This model gives us a kappa statistic of 0.94, which is also close to
what we observed with radial basis function SVM.

```{r}
fit$results
```

Let's again see what type of digits are being misclassified:

```{r}
tab <- table(obs=fit$pred$obs, pred=fit$pred$pred)
prop <- tab/rowSums(tab)
round(prop, 3) * 100 # percent
```

Again, predicting 9 instead of 4 is the worst error, and the reverse
is the second worst error, but the rate has dropped a lot compare to
kNN. 

# Convolutional neural network

Finally, we show an example of a convolutional neural network (CNN)
code using the *keras* R package, which provides an interface to the
Keras Python Deep Learning library. The code that we used below is
copied from the following example. We do not motivate all of the
choices, as they have been tuned specifically for this problem by
researchers that have focused deeply on the MNIST data, handwriting,
and image recognition in general.

* <https://keras.rstudio.com/articles/examples/mnist_cnn.html>

We being by converting the target `y` to a 10 dimensional binary matrix:

```{r}
library(keras)
y_train <- to_categorical(y)
head(y_train)
```

The following parameters are chosen for the MNIST data in particular
(the image size and number of classes are determined obviously by the
problem). 

```{r}
batch_size <- 128
num_classes <- 10
epochs <- 12
img_rows <- 28
img_cols <- 28
```

We reshape `x` as well into an array that takes into account the fact
that we have a 2D array of pixels.

```{r}
x <- as.matrix(dat)
x_train <- array_reshape(x, c(nrow(x), img_rows, img_cols, 1))
dim(x_train)
input_shape <- c(img_rows, img_cols, 1)
```

The following code chunk specifies the model: two 2D convolutional
layers, a 2D pooling layer, a dropout layer, followed by a dense layer
with 28 nodes and ReLU activation function, followed by a dropout
layer, and finally a dense output layer using softmax activation
function. Each *epoch* takes about 20 seconds, and we run 12 epochs in
total. 

```{r}
set.seed(1)

model <- keras_model_sequential() %>%
  layer_conv_2d(filters = 32, kernel_size = c(3,3), activation = 'relu',
                input_shape = input_shape) %>% 
  layer_conv_2d(filters = 64, kernel_size = c(3,3), activation = 'relu') %>% 
  layer_max_pooling_2d(pool_size = c(2, 2)) %>% 
  layer_dropout(rate = 0.25) %>% 
  layer_flatten() %>% 
  layer_dense(units = 128, activation = 'relu') %>% 
  layer_dropout(rate = 0.5) %>% 
  layer_dense(units = num_classes, activation = 'softmax')

# Compile model
model %>% compile(
  loss = loss_categorical_crossentropy,
  optimizer = optimizer_adadelta(),
  metrics = c('accuracy')
  )

# Train model
history <- model %>% fit(
  x_train, y_train,
  batch_size = batch_size,
  epochs = epochs,
  validation_split = 0.2
  )
```

We can plot the accuracy of the model over time. Note that the
validation accuracy approaches 98%, which beats our previous
approaches (the MLP had about 94% accuracy).

```{r}
plot(history)
```

We can evaluate the CNN on a test set (we only used the first 10,000
rows to train our model).

```{r}
x2 <- as.matrix(mnist_raw[10001:20000,-1])/255
x_test <- array_reshape(x2, c(nrow(x2), img_rows, img_cols, 1))
y_test <- to_categorical(factor(mnist_raw$X1[10001:20000]))
```

```{r}
scores <- model %>% evaluate(
  x_test, y_test, verbose = 0
  )
scores
```

We can get the predicted classes:

```{r}
pred <- model %>% predict_classes(x_test)
```

We can see we've further reduced our worst misclassification rate to
1.4%.

```{r}
y.test <- factor(mnist_raw$X1[10001:20000])
tab <- table(obs=y.test, pred=pred)
prop <- tab/rowSums(tab)
round(prop, 3) * 100 # percent
```
