---
title: "HW 9 - MCMC"
author: "Naim Rashid"
date: "2/20/2019"
output: html_document
---
 
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Maximization of poisson GLMM from lecture

Lets now maximize the poisson GLMM model given in lecture, now using an MCEM approach.  In a previous HW, you used numerical integration tools to approximate the likelihood for this model, then applied numerical optimization to obtain the estimates of the model parameters.  

Here we wish to use MCEM, where in lecture we have already gone over a similar implementation using a rejection sampler in the E-step.  

For this HW, please use a Metropolis Hastings Random Walk proposal distribution to approximate the Q-function in the E-step.   Specify your proposal distribution.  Write functions implementing the E-step, the M-step, and then write the main code for the MCEM algorithm.  

Feel free to reuse/modify the lecture code to do this.   However, you can implement the M-step and other parts of the EM algorithm however is most convenient to you.  Not required by any means, but it may be helpful from a speed perspective to recode the sampler into Rcpp. 

```{r}
## Solution: place relevant helper functions pertaining to the E step here 

## End Solution


## Solution: place relevant helper functions pertaining to the M step here 

## End Solution


## Solution: place primary code for the MCEM algorithm here, calling functions in the above two sections
## Remember to print your primary results and use the following starting values, and evaluate chain diagnostics for the final model

# set initial parameters
  tol = 10^-5
  maxit = 100
  iter = 0
  eps = 10000
  qfunction = rep(-10000,maxit) # using Qfunction for convergence
  
# starting values, taken from rejection sampling example
  beta = c(1.804, 0.165) 
  s2gamma =  0.000225 + .01 

# Length of chain
  M = 10000

# burn in
  burn.in = 2000

  
    
## End Solution


```
