---
title: "Building R packages"
author: "Michael Love"
date: 10/25/2018
output: html_document
---

# Why put your code into a package?

There are two compelling reasons to move your code from a set of
scripts to a package:

* Versioning
* Documentation

During a project, you may begin to accumulate a large code base, you
can get by organizing functions into files, and using `source`, for
example: 

```{r eval=FALSE}
# hypothetical example:
load("data.rda")
source("normalize_functions.R")
source("EM_functions.R")
source("plot_functions.R")
```

This may be sufficient for a while, but it is not sufficient for
sharing the code with others, either for reproducibility or for having
others use your methods. Using `source` as above doesn't let you or
someone else know which *version* of the functions are being used. The
way to do this in R is to put the R scripts into a *package structure*
which has a version attached to it. The version is simply recorded as
a line in a text file, which we will describe in these notes.

A potential user also cannot find out about the purpose of the
functions, details about the arguments or output of the function if
they are simply brought into R with `source`. You could provide
comments in the R files, but these comments are not accessible from
the command line while a potential user is about to use your
function. Help for R functions is provided in a specifically formatted
file (ending with `.Rd`) that goes into the `man` directory of an R
package. I strongly recommend to not write these files by hand but
instead use in-line documentation called *roxygen* and help functions
from the *devtools* package to automatically create these help
files. This will be covered in detail in another lecture note.

Two additional references for building R packages are:

* [R Packages](http://r-pkgs.had.co.nz/) by Hadley Wickham (currently
  out of date with respect to the addition of the *usethis* package)
* [Writing R Extensions](https://cran.r-project.org/doc/manuals/R-exts.html)
  the official guide from CRAN on how to create an R package

# Minimal package skeleton

The bare minimum for an R package is a directory with the following:

* a `DESCRIPTION` text file, which gives information about the package
* a `NAMESPACE` text file, which tells which functions are imported or
  exported by the package (best written automatically, not manually)
* a directory called `R` with the R scripts inside

We can get started by using the `create_package()` function from the *usethis*
package. Suppose we have an R script with the following function:

```{r eval=FALSE}
add <- function(x,y,negative=FALSE) {
  z <- x + y
  if (negative) {
    z <- -1 * z
  }
  z
}
```

Let's say we will create an R package called *foo*. We can start with
the following:

```{r eval=FALSE}
library(usethis)
create_package("foo", rstudio=FALSE)
```

This will make a directory called `foo` in the current working
directory, and will then change the working directory to `foo`.
(This Rmarkdown however needs to point to the `foo` directory to look
inside.)

```{r}
list.files("foo")
```

It will start with a DESCRIPTION file that looks like:

```
Package: foobar
Title: What the Package Does (One Line, Title Case)
Version: 0.0.0.9000
Authors@R:
    person(given = "First",
           family = "Last",
           role = c("aut", "cre"),
           email = "first.last@example.com")
Description: What the package does (one paragraph).
License: What license it uses
Encoding: UTF-8
LazyData: true
```

The DESCRIPTION file has a number of fields that we need to
populate. We may start editing this file: 

```
Package: foo
Title: Functions for doing great things
Version: 0.0.1
Authors@R: person("Jane","Doe",email="jdoe@uni.edu",role=c("aut","cre"))
Description: Contains amazing functions for doing amazing things.
License: GPL-2
Encoding: UTF-8
LazyData: true
```

There are a number of licenses that are possible and these are
important to read about and consider before releasing code. Here is a
list of licenses that are in use for R:

<https://www.r-project.org/Licenses/>

R itself is licensed under `GPL-2 | GPL-3`.

If we simply move the R script above into the `R` directory, we are
done with the bare minimum for our first R package, because the
`NAMESPACE` file already says to export all functions that do not
begin with a period:

```
exportPattern("^[^\\.]")
```

A complication is if we want to use functions from another package in
our package. The way to do this is to *import* or *depend* on other
packages. The easiest way to handle importing functions from another package
is through the *roxygen* documentation format, so we will hold off on
discussing this in the next lecture note. We have some description of
the different kinds of package dependencies at the bottom of this
lecture note.

We can build a shareable package "tarball" using the `build` function:

```{r eval=FALSE}
build("foo")
```

Alternatively, this can be done on the command line, from one
directory above the package directory, with `R CMD build foo`.

The `build` command prints out the following message and returns the
location of the package tarball.

```
'/Users/love/bin/R/bin/R' --no-site-file --no-environ --no-save --no-restore --quiet CMD build  \
  '/Users/love/teach/statcomp/statcomp_src/rpkg/foo' --no-resave-data --no-manual 

* checking for file ‘/Users/love/teach/statcomp/statcomp_src/rpkg/foo/DESCRIPTION’ ... OK
* preparing ‘foo’:
* checking DESCRIPTION meta-information ... OK
* checking for LF line-endings in source and make files and shell scripts
* checking for empty or unneeded directories
* building ‘foo_0.0.1.tar.gz’

[1] "/Users/love/teach/statcomp/statcomp_src/rpkg/foo_0.0.1.tar.gz"
```

Now if we want to share our function `add` with a collaborator, we can
send them the file `foo_0.0.1.tar.gz`. 

# Version numbers

The most important thing about version numbers is that they are
*free*. You should "bump" the version number any time you make a
change that you will push out the world (e.g. anytime you push your
changes to GitHub or other repository). Even if you are working
entirely by yourself, it is useful to be able to pinpoint differences
in code output based on the version string (which will appear if you
append the session information to the output of every script or
Rmarkdown file you run).

There are three parts to most R package version numbers:

`x.y.z`

Roughly:

* The "x" part is for major releases
* The "y" part is for minor releases
* The "z" part is for any change to the package

In *Bioconductor*, there is some additional information and restrictions
on these numbers. "y" is even for release versions and odd for
development versions. They have additional information on the system
for [Bioconductor package version numbers](http://bioconductor.org/developers/how-to/version-numbering/).

The first time a package is started, you may choose `0.0.1` and then
add 1 to the "z" digit every time you make a change to the package
code. It would be typical to use `1.0.0` when you first "release" the
package, which you would do yourself if you are planning to host on
GitHub or on CRAN. If you still consider the package a working "beta",
then you might only use `0.1.0`, until you feel it is ready for
`1.0.0`. For Bioconductor packages, you would submit
`0.99.z` to the Bioconductor repository, and the Bioconductor machines
would increment to `1.0.0` on your behalf for the first release.
There is some additional information about version numbers and
releasing a package from [Hadley Wickham](http://r-pkgs.had.co.nz/release.html).

In R, package versions are a special class and you can compare package
versions like so: 

```{r}
packageVersion("stats")
packageVersion("stats") >= "3.0.0"
```

If you want a R script to produce an error if the package is out of
data you can use:

```{r}
stopifnot(packageVersion("stats") >= "3.0.0")
```

For an R package, if you need a specific version of another package,
you should include a string in parentheses after its name in the
Depends, Imports or Suggests field, for example:

```
Imports: foo (>= 1.2.0)
```

We will discuss these fields a bit more in the documentation notes.

# Loading and sharing packages

By far the easiest way to load the package into R while you are
developing it is to use the `load_all` function from the *devtools*
package. This does not require the creation of the package tarball
with the `.tar.gz` ending, but simply mimics an installation of the
package. From within the package you can simply call `load_all()`, or 
you can also specify the path to the package:

```{r eval=FALSE} 
load_all("/path/to/foo")
```

You also bypass having to call `library(foo)` as the package will
already be loaded.

Another way to load the package is the standard `install.packages`
function but specifying that the package tarball is a local file, not
a name of a package in a remote repository. This then requires an
explicit `library(foo)` call afterward.

```{r eval=FALSE}
install.packages("foo_0.0.1.tar.gz", repos=NULL)
```

Another easy way to share your package is to put all of the files into
a GitHub repository. Then others can install the package on their
machines simply with `install_github("username/foo")` using the
*devtools* package. Again, this requires a `library(foo)` call
afterward to load the package.

We still haven't written any documentation, so when we try to load
help we are told it is missing:

```{r}
library(foo)
help(topic="add", package="foo")
```

Either through using `load_all` or installing and then loading with
`library`, we can now use our function:

```{r}
add(3,4)
add(3,4,negative=TRUE)
```

And for reproducibility sake, we can ask about the package version:

```{r}
packageVersion("foo")
```

And we can include the following at the end of scripts to include all
package versions (and other important details):

```{r}
session_info()
```

# Types of package dependencies

There are three main types of formal package dependencies in R:

* Depends
* Imports
* Suggests

Each of these can be listed in the `DESCRIPTION` file. We will show in
the documentation lecture note how to import specific functions from
other packages, and how to add these packages to the `Imports:` line
in the `DESCRIPTION` file. I recommend to import specific functions
from other packages, rather than the entire package, but it is also
possible to import the entire package. Note that the packages that
you import functions from are not *attached to the search path* when
you load your package with `library(foo)`. This means that, even if
you import one or more functions from another package within your
package, those functions are not available to the user unless they
also load the package with another call to `library`.

The `Suggests:` field of the `DESCRIPTION` file is for packages which
are not absolutely required for your package to work, but which your
package can make use of, or which are used in examples or the vignette
of your package. I recommend to list a package under Suggests also in
the case that you have a function which can make use of another
package, but has a fallback implementation which does not require the
other package. Or if there is no fallback, the function should explain
that it cannot be run because the other package is not available. This
testing to see if the other package is available can be performed using
`requireNamespace("foo", quiet=TRUE)`, which will return `TRUE` if the
package is installed and `FALSE` if not. You can then use the function
from the other package with `::` within your package, for example:
`cool::fancy` can be used to call the `fancy` function from the *cool*
package.

This leaves Depends to be explained. If you read Hadley Wickham's
R Package reference and particular the section on
[Namespace](http://r-pkgs.had.co.nz/namespace.html), you'll see that
he does not recommend the use of Depends at all. I agree with his
arguments laid out there and simply quote the key sentences here:

> The main difference is that where Imports just loads the package,
> Depends attaches it. There are no other differences.
> ... Unless there is a good reason otherwise, you
> should always list packages in Imports not Depends. That’s because a
> good package is self-contained, and minimises changes to the global
> environment (including the search path).

# Package checking

R has a very useful, but also very picky checking software associated
with it. It can be run with the `check` function from the *devtools*
package, or by running `R CMD check foo_0.0.1.tar.gz` from the command
line. Passing all of the R package checks is a good idea if you plan
to share your code with others. In addition, there may be 
[other checks](https://bioconductor.org/packages/devel/bioc/vignettes/BiocCheck/inst/doc/BiocCheck.html) 
that you will need to pass if you want to submit to Bioconductor. We
won't cover R package checking, as we only have limited time, but the
reference on [R Packages](http://r-pkgs.had.co.nz/check.html) from
Hadley Wickham discusses this.

