---
title: "Building R packages"
author: "Michael Love"
date: 10/25/2018
output: html_document
---

# Why put your code into a package?

There are two compelling reasons to move your code from a set of
scripts to a package:

1) Versioning
2) Documentation

During a project, you may begin to accumulate a large code base, you
can get by organizing functions into files, and using `source`, for
example: 

```{r eval=FALSE}
# hypothetical example:
load("data.rda")
source("normalize_functions.R")
source("EM_functions.R")
source("plot_functions.R")
```

This may be sufficient for a while, but it is not sufficient for
sharing the code with others, either for reproducibility or for having
others use your methods. Using `source` as above doesn't let you or
someone else know which *version* of the functions are being used. The
way to do this in R is to put the R scripts into a *package structure*
which has a version attached to it. The version is simply recorded as
a line in a text file, which we will describe in these notes.

A potential user also cannot find out about the purpose of the
functions, details about the arguments or output of the function if
they are simply brought into R with `source`. You could provide
comments in the R files, but these comments are not accessible from
the command line while a potential user is about to use your
function. Help for R functions is provided in a specifically formatted
file (ending with `.Rd`) that goes into the `man` directory of an R
package. I strongly recommend to not write these files by hand but
instead use in-line documentation called *roxygen* and help functions
from the *devtools* package to automatically create these help
files. This will be covered in detail in another lecture note.

# Minimal package skeleton

The bare minimum for an R package is a directory with the following:

* a `DESCRIPTION` text file, which gives information about the package
* a `NAMESPACE` text file, which tells which functions are imported or
  exported by the package (best written automatically, not manually)
* a directory called `R` with the R scripts inside

We can get started by using the `create` function from the *devtools*
package. Suppose we have an R script with the following function:

```{r eval=FALSE}
add <- function(x,y,negative=FALSE) {
  z <- x + y
  if (negative) {
    z <- -1 * z
  }
  z
}
```

Let's say we will create an R package called *foo*. We can start with
the following:

```{r eval=FALSE}
create("foo", rstudio=FALSE)
```

This will make a directory called `foo` in the current working
directory:

```{r}
list.files("foo")
```

It will start with a DESCRIPTION file that looks like:

```
## Package: foo
## Title: What the Package Does (one line, title case)
## Version: 0.0.0.9000
## Authors@R: person("First", "Last", email = "first.last@example.com", role = c("aut", "cre"))
## Description: What the package does (one paragraph).
## Depends: R (>= 3.6.0)
## License: What license is it under?
## Encoding: UTF-8
## LazyData: true
```

The DESCRIPTION file has a number of fields that we need to
populate. We may start editing this file: 

```
Package: foo
Title: Functions for doing great things
Version: 0.0.1
Authors@R: person("Jane","Doe",email="jdoe@uni.edu",role=c("aut","cre"))
Description: Contains amazing functions for doing amazing things.
Depends: R (>= 3.6.0)
License: GPL-2
Encoding: UTF-8
LazyData: true
```

There are a number of licenses that are possible and these are
important to read about and consider before releasing code. Here is a
list of licenses that are in use for R:

<https://www.r-project.org/Licenses/>

R itself is licensed under `GPL-2 | GPL-3`.

If we simply move the R script above into the `R` directory, we are
done with the bare minimum for our first R package, because the
`NAMESPACE` file already says to export all functions that do not
begin with a period:

```
exportPattern("^[^\\.]")
```

A complication is if we want to use functions from another package in
our package. The way to do this is to *import* or *depend* on other
packages. The easiest way to handle these imports and dependencies is
through the *roxygen* documentation format, so we will hold off on
discussing this in the next lecture note.

We can build a shareable package "tarball" using the `build` function:

```{r eval=FALSE}
build("foo")
```

This prints out the following message and returns the location of the
package tarball.

```
'/Users/love/bin/R/bin/R' --no-site-file --no-environ --no-save --no-restore --quiet CMD build  \
  '/Users/love/teach/statcomp/statcomp_src/rpkg/foo' --no-resave-data --no-manual 

* checking for file ‘/Users/love/teach/statcomp/statcomp_src/rpkg/foo/DESCRIPTION’ ... OK
* preparing ‘foo’:
* checking DESCRIPTION meta-information ... OK
* checking for LF line-endings in source and make files and shell scripts
* checking for empty or unneeded directories
* building ‘foo_0.0.1.tar.gz’

[1] "/Users/love/teach/statcomp/statcomp_src/rpkg/foo_0.0.1.tar.gz"
```

Now if we want to share our function `add` with a collaborator, we can
send them the file `foo_0.0.1.tar.gz`. 

We can install this package using the following:

```{r eval=FALSE}
install.packages("foo_0.0.1.tar.gz", repos=NULL)
```

We still haven't written any documentation, so when we try to load
help we are told it is missing:

```{r}
library(foo)
help(topic="add", package="foo")
```

But we can now use our function:

```{r}
add(3,4)
add(3,4,negative=TRUE)
```

And for reproducibility sake, we can ask about the package version:

```{r}
packageVersion("foo")
```

And we can include the following at the end of scripts to include all
package versions (and other important details):

```{r}
session_info()
```
