---
title: "Debugging in R"
author: "Michael Love"
date: 10/25/2018
output: html_document
---

# Many approaches to debug code in R

There are a number of different ways to do debugging of R code, but as
with documentation and testing, I will show here what I think is the
simplest way to debug a complex situation. Suppose we have a function
that is called in the middle of a larger operation and we want to
inspect its arguments and step through what is happening. My preferred
method for this is to insert the `browser()` command into the function
at a relevant level, and then to run the function in the setting that
produces the behavior I want to investigate. Note that this approach
can also be used in general, not just for debugging, but for when
introducing new features that may alter core components of your
current package infrastructure. 

If I want to investigate an error or behavior in a package that I do
not maintain, I will typically obtain the latest source code for the
package, and apply this same procedure of inserting `browser()` at
critical points that I want to investigate. `load_all()` can be run to
load the version of the package that contains the `browser()` call
instead of the original package source.

I will now create a series of inter-dependent functions, so that we
can examine how different strategies of debugging look in R.

```{r}
# a high level function
high <- function(obj) {
  obj.sub <- subset(obj)
  obj.sub <- analyze(obj.sub)
  return(obj.sub)
}
# some subsetting operation
subset <- function(obj) {
  # every other element
  idx <- seq_len(length(obj$x)/2)*2 - 1
  list(x=obj$x[idx],
       y=obj$y[idx],
       foo=obj$foo)
}
# another sub-function
analyze <- function(obj) {
  if (obj$foo == "no") {
    z <- obj$x + obj$y
  } else if (obj$foo == "yes") {
    z <- obj$x + obj$y + special(obj$x)
  }
  obj$z <- z
  obj
}
# a low-level function
special <- function(x) {
  sum(x) + "hi"
}
```

If we run this with `foo="yes"` we will run into some problems.

```{r error=TRUE}
obj <- list(x=1:10,
            y=11:20,
            foo="yes")
high(obj)
```

We can call `traceback()` which gives the series of function calls
that produced the latest error:

```{r eval=FALSE}
traceback()
```

This gives:

```
3: special(obj$x) at #5
2: analyze(obj.sub) at #3
1: high(obj)
```

My favorite strategy is to put `browser()` inside of one of the
functions, for example, here:

```{r eval=FALSE}
special <- function(x) {
  browser()
  sum(x) + "hi"
}
```

Then rerunning:

```{r eval=FALSE}
high(obj)
```

We are given a special `Browse` prompt where we can see what is in the
environment with `ls()`, and investigate the contents:

```
Called from: special(obj$x)
Browse[1]> 
debug at #3: sum(x) + "hi"
Browse[2]> ls()
[1] "x"
Browse[2]> x
[1] 1 3 5 7 9
Browse[2]> sum(x)
[1] 25
Browse[2]> 
```

To exit, type `Q`.

Another option to debug on error is to set a global option:

```{r eval=FALSE}
options(error=recover)
```

Then when running a function that triggers an error, R will launch a
special browser prompt, but first asks us what level of the stack of
calls we would like to enter:

```
Error in sum(x) + "hi" (from #2) : non-numeric argument to binary operator
Calls: high -> analyze -> special

Enter a frame number, or 0 to exit   

1: high(obj)
2: #3: analyze(obj.sub)
3: #5: special(obj$x)

Selection: 3
```

Here we pick to enter frame number 3.

```
Called from: analyze(obj.sub)
Browse[1]> ls()
[1] "x"
Browse[1]> x
[1] 1 3 5 7 9
Browse[1]> sum(x)
[1] 25
```

When you are finished, you can return with `Q` and you can stop the
browser prompt upon error with:

```{r eval=FALSE}
options(error=NULL)
```

Other options for debugging in R are the `debug`/`undebug` or `debugonce`
functions. These are similar to `browser` but they step through a function
line-by-line, and so can be tedious for large functions. However they
do not require having the source code of the program on hand for
editing.
