---
title: "Documenting an R package"
author: "Michael Love"
date: 10/25/2018
output: html_document
---

# Types of documentation in R

There are two main locations in an R package which contain
documentation:

* help pages (sometimes called "man pages") for each function
* vignettes (longer form with R code chunks embedded in text)

We will show in this lecture note how to generate documentation for
both. For the function help pages, I strongly recommend to use
*roxygen* formatting and the `document` function from the *devtools*
package, which will be covered here. This will automatically generate
both the `.Rd` files that go in the `man` directory, and it will
populate the `NAMESPACE` file which specifies what is exported and
what is imported by your package.

# Writing help for functions

Suppose we have a function in one of our R scripts, which is located
in the `R` directory of our package.

```{r}
add <- function(x,y,negative=FALSE) {
  z <- x + y
  if (negative) {
    z <- -1 * z
  }
  z
}
```

We need to tell our prospective users what the function does, what
kind of arguments it takes, and what is the output of the function. It
is useful to also supply references to literature or to specific
manuscripts associated with the code.

Take a look at the help for the `quantile` function and the `sapply`
function: 

```{r eval=FALSE}
?quantile
?sapply
```

You can see there is some variation among function help pages, but key
elements are repeated. Here we will cover some basics about
documenting functions, and for further reference you can look up
the notes from [Hadley Wickham](http://r-pkgs.had.co.nz/man.html) on
writing documentation.

# roxygen

To start documenting this function with *roxygen*, you type a pound
key and a single quote mark `#'` in a line above the function, and
then add a title to the function

```
#' Sum of two vectors of numbers
add <- function(x,y,negative=FALSE) {
...
```

If you press Enter for a new line from the title line, in most R-aware
editors (Rstudio, Emacs + ESS, although not the R GUI) you will get a
new line starting with `#'`.

The first line becomes the Title of the help page, the second line
becomes the Description (briefly describe what the function does), and
further lines will populate the Details section of the help page. For
example:

```
#' Sum of two vectors of numbers
#' 
#' This function sums two vectors of numbers and optionally
#' allows for the negative of the sum to be returned.
#' 
#' This paragraph will start the Details section...
#' 
#' This will be the second paragraph of the Details section... 
add <- function(x,y,negative=FALSE) {
...
```

If it is desired to refer to the function itself or another function,
one can use the following, `\code{add}`, which will be rendered as
monospace type, `add`, in the HTML and PDF version of the help pages and
as `'add'` with single-quotes in the R console version of the help pages.

# Arguments

Next we will document the arguments, which get a special tag to
indicate their lines `@param`. One may ask why the arguments are
tagged with `@param` as they will show up in the help page under
**Arguments** and I believe the answer is that *roxygen* for
documenting R code is patterned historically on
[Doxygen](https://en.wikipedia.org/wiki/Doxygen) which uses the tag
`@param`.

Here we take out the extra Details paragraphs from above, and just
focus on the Title, Description and Arguments:

```
#' Sum of two vectors of numbers
#' 
#' This function sums two vectors of numbers and optionally
#' allows for the negative of the sum to be returned.
#' 
#' @param x a vector of numbers
#' @param y a vector of numbers
#' @param negative logical, whether to flip the sign of the sum
add <- function(x,y,negative=FALSE) {
...
```

The format is: `@param name description`. The description is a bit
personal preference, and I tend to put the expected type of the
argument (e.g. logical) in the front for certain arguments, and
sometimes also the default value: "logical, whether to flip the sign
of the sum (default is FALSE)". The default value will also be printed
in the Usage section, which is generated by default, so it's not
strictly necessary.

# Returned values

It's also important to add the Value that is returned by the
function, which I tend to put below the `@param` lines. In this
trivial example, it's not very revealing, but some functions have
complex outputs, e.g. a list with different elements, or a complex
object, in which case it is useful to describe exactly what is being
returned. If there is any ambiguity about the returned values,
*please* indicate it in the help file, for example if the values are
on the log, log2 or log10 scale, this would be the place to describe
it. 

```
...
#' @param negative logical, whether to flip the sign of the sum
#' 
#' @return the sum of the two vectors
add <- function(x,y,negative=FALSE) {
...
```

If a list-like object is being returned, one can use the following
paradigm to describe each piece of the returned object:

```
#' @return a list with the following elements:
#' \itemize{
#' \item{...}
#' \item{...}
#' }
```

This will become a bulleted list in the help page. This paradigm can
be used in the other sections, e.g. arguments as well.

# Examples

In the Examples field, you can provide small examples of using the
function *that will run when you check the package*. These are
required by Bioconductor. Ideally the examples should take no more
than a few seconds. The R code directly follows the line with the
tag `@examples`:

```
...
#' @return the sum of the two vectors
#' 
#' @examples
#'
#' add(1:5, 6:10)
#' add(1:5, 6:10, negative=TRUE)
#'
add <- function(x,y,negative=FALSE) {
...
```

# Import and export

While *roxygen* is not the only way to deal with function import and
export, I find it the easiest by far. Any function which you desire to
export (to make visible to users who load your package), you add the
`@export` tag in the roxygen code. I tend to add it directly above the
function definition:

```
...
#' @export
add <- function(x,y,negative=FALSE) {
...
```

Imports are only slightly more complicated. If you want to import a
specific function from a package (and I recommend this, rather than
importing entire packages), you can add the `@importFrom` tag in the
*roxygen* code block. The format is `@importFrom package function`. I
also recommend to use `package::function` in your R code when you use
a function from another package. This clarifies that the function is
not defined in your package and will avoid errors when you run R's
package check.

```
...
#' @importFrom gtools rdirichlet
#' 
#' @export
add <- function(x,y,negative=FALSE) {
  d <- gtools::rdirichlet(1, alpha=c(1,2,3))
  z <- x + y
  ...
```

# Actually making the Rd files

Now that we have created the following *roxygen* code block above our
file, we can run the `document` function from the *devtools* package
to create the `.Rd` files for our package.

Altogether we have:

```
#' Sum of two vectors of numbers
#' 
#' This function sums two vectors of numbers and optionally
#' allows for the negative of the sum to be returned.
#' 
#' @param x a vector of numbers
#' @param y a vector of numbers
#' @param negative logical, whether to flip the sign of the sum
#'
#' @return the sum of the two vectors
#' 
#' @examples
#'
#' add(1:5, 6:10)
#' add(1:5, 6:10, negative=TRUE)
#'
#' @importFrom gtools rdirichlet
#' 
#' @export
add <- function(x,y,negative=FALSE) {
  d <- gtools::rdirichlet(1, alpha=c(1,2,3))
  z <- x + y
  if (negative) {
    z <- -1 * z
  }
  z
}
```

When we run `document()` the help pages with `.Rd` ending are created
and also the `NAMESPACE` file is re-written to indicate the imports and
exports. The `DESCRIPTION` file may also be updated with a single
field `RoxygenNote: x.y.z` denoting the version of *roxygen2*, the R
package that handles reading *roxygen* code chunks and writing `.Rd`
files. 

The following file is created in `man/add.Rd`. Note that it says the
file is generated by the *roxygen2* package and so **should not be
edited by hand**. This is because the source of the help page is in
the R file, and so any edits to this file will be wiped out the next
time `document()` is run.

```
% Generated by roxygen2: do not edit by hand
% Please edit documentation in R/foo.R
\name{add}
\alias{add}
\title{Sum of two vectors of numbers}
\usage{
add(x, y, negative = FALSE)
}
\arguments{
\item{x}{a vector of numbers}

\item{y}{a vector of numbers}

\item{negative}{logical, whether to flip the sign of the sum}
}
\value{
the sum of the two vectors
}
\description{
This function sums two vectors of numbers and optionally
allows for the negative of the sum to be returned.
}
\examples{

add(1:5, 6:10)
add(1:5, 6:10, negative=TRUE)

}
```

Note that, if you are keeping your package on a repository such as
GitHub, you will need to explicitly add `man/foo.Rd` to the repository
and push it to the origin, so that others will have access to the
documentation when they load your package.

Another note: if you just want to preview the help file, you can
simply call `load_all()` and then type out `?foo`, which will bring up
your most recent edits.

The NAMESPACE file is updated from exporting all functions that do not
begin with a period, to only exporting the `add` function:

```
# Generated by roxygen2: do not edit by hand

export(add)
importFrom(gtools,rdirichlet)
```

There is one thing left to do to manually, which is to add the
following line to the `DESCRIPTION` file, which indicates that we are
importing at least one function from the *gtools* package:

```
Imports: gtools
```

# Vignettes


```
---
title: "A great title for a vignette"
date: "`r format(Sys.Date(), '%m/%d/%Y')`"
author: "Jane Doe, John Doe"
output: rmarkdown::html_document
abstract: |
  Lorem ipsum lorem ipsum lorem ipsum lorem ipsum lorem ipsum
  lorem ipsum lorem ipsum lorem ipsum lorem ipsum lorem ipsum
bibliography: library.bib
vignette: |
  %\VignetteIndexEntry{A great title for a vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
```

Add *knitr* and *rmarkdown* to Suggests...

```
VignetteBuilder: knitr, rmarkdown
```
