---
title: "Homework 3 - C++ code from R"
author: "your name here"
date: "`r format(Sys.time(), '%m/%d/%Y')`"
output: html_document
---

# Setup and hints

For all of the following functions, you should add these functions to
your R package, by adding code to the `src` directory, adding the
necessary importing and linking to lines to `DESCRIPTION`, and then
writing and exporting a wrapper R function (e.g. adding a `@export`
tag). 

Remember to include the following at the top of your C++ script (you
do not include `Rcpp.h` because `RcppArmadillo.h` includes this). As
the homework uses RcppArmadillo, the following covers all questions:

```
#include <RcppArmadillo.h>
using namespace Rcpp;
```

Remember that Armadillo uses 0-based indexing. This is true for square
bracket indexing `z[0]` or `Z[0,0]` as well as using the row and
column getters: `z.row(0)` and `z.col(0)`.

# Question 1 - simple Rcpp function

Write a simple Rcpp function that implements the following R code:

`ifelse(x < 0.0, 1.0, exp(x))`

Compare it's output in R. Your R wrapper function should be named
`one.or.exp()`.

```{r}
library(devtools)
document("bios735")
load_all("bios735")
x <- -10:10
one.or.exp(x)
all.equal(ifelse(x < 0.0,1.0,exp(x)), one.or.exp(x))
```

# Question 2 - random walk in C++ and vectorized R

Write a random walk function in C++, call the wrapper function for
your C++ function, `randomWalk2`. Your C++ function should follow the
same style as the function below, by iterating over 1-niter, despite
the fact that the random walk can be vectorized in R as a cumulative
sum. You will later compare to the vectorized random walk in R.

Hint: you should use the C++ function `Rf_rbinom(integer size, numeric
probability)` function to flip the coin in the random walk.

Compare the output of `randomWalk2` to ensure that it produces the
same output as `randomWalk1`, with the same seed set in R. How much
faster is your random walk? 

```{r}
niter <- 1e4
lambda <- .01
randomWalk1 <- function(niter,lambda) {
  x <- numeric(niter)
  y <- numeric(niter)
  for (i in seq_len(niter)[-1]) {
    x[i] <- x[i-1] + lambda * (2.0 * rbinom(1,1,0.5) - 1.0)
  }
  for (i in seq_len(niter)[-1]) {
    y[i] <- y[i-1] + lambda * (2.0 * rbinom(1,1,0.5) - 1.0)
  }
  list(x=x,y=y)
}
myplot <- function(dat) {
  niter <- length(dat$x)
  plot(0,type="n",xlim=c(-2,2),ylim=c(-2,2),xlab="x",ylab="y")
  cols <- colorRampPalette(c("blue","red","orange"))(100)
  with(dat, segments(x[-niter],y[-niter],x[-1],y[-1],col=rep(cols,each=niter/100)))
}
set.seed(5)
dat1 <- randomWalk1(niter,lambda)
str(dat1)
myplot(dat1)
```

Your random walk:

```{r}
document("bios375")
load_all("bios375")
set.seed(5)
dat2 <- randomWalk2(niter,lambda)
str(dat2)
myplot(dat2)
all.equal(dat2$x, dat1$x)
all.equal(dat2$y, dat1$y)
library(microbenchmark)
microbenchmark(randomWalk1(niter,lambda),randomWalk2(niter,lambda),times=10)
```

Now write a vectorized version of the random walk in R and call this
function `randomWalkVectorized`. Compare to ensure it gives the same
result at `randomWalk1`, and compare its speed to `randomWalk1` and to
`randomWalk2`:

```{r}
randomWalkVectorized <- function(niter,lambda) {
  # your code here...
}
set.seed(5)
datVec <- randomWalkVectorized(niter,lambda)
str(datVec)
myplot(datVec)
all.equal(datVec$x, dat1$x)
all.equal(datVec$y, dat1$y)
library(microbenchmark)
microbenchmark(randomWalk1(niter,lambda),randomWalkVectorized(niter,lambda),times=10)
microbenchmark(randomWalk2(niter,lambda),randomWalkVectorized(niter,lambda),times=10)
```

# Question 3 - simple RcppArmadillo function

Write a simple RcppArmadillo function that solves for `x` in the
matrix multiplication formula $Ax = b$. Call your R wrapper function
`armadilloSolve`. You can skip the part about pointing to the matrix
in R with `const arma::mat& A` for this and the following homework
question. That is, the top line of your function can look like:

```
arma::mat armadillo_solve(arma::mat A, arma::vec b) {
```

Check that your function gives the correct answer here:

```{r}
document("bios735")
load_all("bios735")
A <- matrix(runif(12),ncol=3)
x <- matrix(runif(3),ncol=1)
b <- A %*% x
xx <- armadilloSolve(A,b)
all.equal(x, xx)
```

# Question 4 - column-wise ridge regression

Implement a C++ function that performs ridge regression one-by-one on
columns of a matrix `Y`, using a constant design matrix `X`, and a
variable ridge parameter `lambda`. That is, use the first element of
`lambda` for the first column of `Y` and so on. Call your R wrapper
function `colRidge2`. Again, you can skip the part about pointing from
C++ to the matrix in R and just have the top of your C++ function look
like, e.g.

```
arma::mat col_ridge_2(arma::mat Y, arma::mat X, arma::vec lambda) {
```

Compare the output of `colRidge2` to `colRidge1`. It is not expected to
be numerically identical, but it should be "close" which you can see
from the plots. How much faster is the C++ version?

Hint: for your C++ function use the simple ridge formula:

$\hat{\beta} = (X^t X + \lambda I)^{-1} X^t y$

Hint: see the `eye()` function in Armadillo to build the identity matrix.

```{r}
document("bios735")
load_all("bios735")
set.seed(1)
n <- 100
Y <- matrix(rnorm(n*20),nrow=20)
X <- scale(matrix(rnorm(20*2),ncol=2))
lambda <- runif(n,.1,2)
library(MASS)
colRidge1 <- function(Y, X, lambda) {
  df <- as.data.frame(X)
  n <- ncol(Y)
  beta <- matrix(nrow=2,ncol=n)
  stopifnot(length(lambda) == n)
  for (j in seq_len(n)) {
    beta[,j] <- coef(lm.ridge(Y[,j] ~ 0 + V1 + V2, data=df, lambda=lambda[j]))
  }
  beta
}
beta1 <- colRidge1(Y, X, lambda)
beta1[,1:5]
beta2 <- colRidge2(Y, X, lambda)
beta2[,1:5]
plot(beta1[1,], beta2[1,])
abline(0,1)
plot(beta1[2,], beta2[2,])
abline(0,1)
all.equal(beta1[1,], beta2[1,])
all.equal(beta1[2,], beta2[2,])
microbenchmark(colRidge1(Y, X, lambda), colRidge2(Y, X, lambda), times=10)
```
