---
title: "Writing package tests"
author: "Michael Love"
date: 10/25/2018
output: 
  html_document:
    toc: true
    toc_float: true
---

# Why write tests?

Package tests are a simple way to make sure that the statistical
software you have written does what you expect, both when you run it
on typical and atypical input. I also tend to use package tests when
implementing new features in my software packages, as a way to check
to see that the new functionality works as I expect it.

There are multiple frameworks for writing package tests, but we will
focus on the framework that I find the most straightforward, which is
implemented in the *testthat* package.

A reference for writing tests with *testthat* can be found at the [R
Packages](http://r-pkgs.had.co.nz/tests.html) book by Hadley Wickham.

# Set up testthat for a package

To begin writing tests, say for a part of your software that you call
"name", you can run the *usethis* function `use_test("name")`. This
will create a directory called `tests/testthat` in the root of your R
package directory, add *testthat* to your `Suggests:` line in the
`DESCRIPTION` file, create a file `tests/testthat.R` that will run
all the tests in `tests/testthat` when you run R's package check, and
create a file `tests/testthat/test-name.R`. You may have multiple
groups of tests that you want to separate into different files, so you
can choose "name" however you like, e.g. `test-data-input.R`,
`test-normalization.R`, etc. However, you can also put all your tests
into a single file for the package, e.g. `test-foo.R`.

The `testthat.R` file is very simple:

```
# This file is part of the standard setup for testthat.
# It is recommended that you do not modify it.
#
# Where should you do additional test configuration?
# Learn more about the roles of various files in:
# * https://r-pkgs.org/tests.html
# * https://testthat.r-lib.org/reference/test_package.html#special-files

library(testthat)
library(foo)

test_check("foo")
```

This file stays the same way, and we will write new `.R` files that go
into `tests/testthat` which will implement the package tests.

Suppose we run `use_test("add")` for our *foo* package, and we want to
write a test for our `add` function (make sure the *usethis* pakage is loaded). We can do this by opening up the
file `tests/testthat/test-add.R`, and adding some tests. The default
file has some dummy code to show you the style:

```
test_that("multiplication works", {
  expect_equal(2 * 2, 4)
})
```

But we can rewrite this for our purposes:

```
test_that("add works on two vectors", {

  expect_equal(add(1:5,6:10), c(7,9,11,13,15))

})

test_that("simple errors for bad input", {

  expect_error(add())
  expect_error(add(1:5))
  expect_error(add(1:5,6:10,"yes"))

})
```

There are many possible tests that one can write, with the workhorses
probably being `expect_equal` and `expect_true`. We can also specify a
numerical tolerance (absolute or relative) for equality, as shown in
the Examples in `?expect_equal`. In order to see a list of all the
`expect_` functions available in *testthat*, one can run the following
command in R:

```{r eval=FALSE}
help(package="testthat", help_type="html")
```

# Messages, warnings, and errors

We can also check that specific messages, warnings, or errors are
output for given input to our function. These three levels of output
`message` the user relevant information, provide a `warning` to the
user about potential problems, or `stop` the function from providing
any output.

If we wanted the `add` function to warn the user about negative values
as output (just a trivial example), we could write:

```{r}
add2 <- function(x,y,negative=FALSE) {
  z <- x + y
  if (negative) {
    z <- -1 * z
  }
  if (any(z < 0)) {
    warning("some output values are negative")
  }
  z
}
```

We could then test this by saying we expect a specific warning. Note
that the entire warning doesn't need to be written out, only a regular
expression that would produce a match.

```{r}
library(testthat)
expect_warning(add2(1:5, -11:-15), "are negative")
```

If we wanted to test for a message or error, we would use
`expect_message` or `expect_error` with the `message` or `stop`
function respectively.

# Testing files or packages

We can check all the tests for individual files with the following
call to `test_file`, from within the package root:

```{r eval=FALSE}
library(devtools)
load_all()
test_file("tests/testthat/test-add.R")
```

Or we can check all of the tests for a given package with the
following call to `test_package`:

```{r eval=FALSE}
test_package("foo")
```

# Session info

```{r}
library(devtools)
session_info()
```
