---
title: "Working with rhdf5"
author: "Michael Love"
date: 11/7/2018
output: html_document
---

In previous lectures, we have discussed reading in large data tables,
and working with large databases via *SQLite*. Here, we discuss a
middle way, using the popular *HDF5* format. The motivation for using
an *HDF5* data container is that, like *SQLite* we have a common
format for representing a complex set of tables that can be shared
simply be sharing a file, but unlike *SQLite* we are typically
interested in reading in entire tables into memory, so that we can
then analyze them. *HDF5* is typically smaller on disk, as well as
faster for writing or reading to or from disk, compared to *SQLite*. 

First some information from the *HDF5* group, on 
[Why HDF5?](https://support.hdfgroup.org/HDF5/faq/whyhdf5.html)

> An HDF5 data container is a standardized, highly-customizable data
> receptacle designed for portability. Unless your definition of
> 'container' is extremely broad, file systems are not commonly
> considered containers. 
>
> File systems aren't portable: For example, you might be able to
> mount an NTFS file system on an AIX machine, but the integers or
> floating point numbers written on an Intel processor will turn out
> to be garbage when read on a IBM Power processor. 
>
> HDF5 achieves portability by separating its "cargo" (data) from its
> environment (file system, processor architecture, etc.) and by
> encoding it in a self-describing file format. The HDF5 library
> serves the dual purpose of being a parser/encoder of this format and
> an API for user-level objects (datasets, groups, attributes, etc.). 
>
> ...
>
> The data stored in HDF5 datasets is shaped and it is typed. Datasets
> have (logically) the shape of multi-dimensional rectilinear
> arrays. All elements in a given dataset are of the same type, and
> HDF5 has one of the most extensive type systems and one that is
> user-extendable.

# The rhdf5 package

As we are focusing on how to interface with various large data formats
in R, we now introduce the *rhdf5* package. Unlike some of the other
packages we have shown, this package is maintained on the Bioconductor
repository and so has a special installation. 

```{r eval=FALSE}
install.packages("BiocManager") # can be skipped after 1st time
BiocManager::install("rhdf5")
```

Now we can load the package. Much of the following introduction to
*rhdf5* is modified from the package vignette.

```{r}
library(rhdf5)
```

Typically, we may already have an *HDF5* data container that we want
to work with, but as in the *SQLite* lecture note, we will show how to
create a new one first.

```{r}
h5file <- "myDB.h5"
h5createFile(h5file)
```

*HDF5* containers have a hierarchy built around *groups* which act and
look a bit like directories:

```{r}
h5createGroup(h5file, "A")
h5createGroup(h5file, "B")
h5createGroup(h5file, "A/C")
```

We can list the groups:

```{r}
h5ls(h5file)
```

Finally, we show some examples of writing data to the *HDF5*
container: 

```{r}
A = matrix(1:10,nr=5,nc=2)
h5write(A, "myhdf5file.h5","foo/A")
B = array(seq(0.1,2.0,by=0.1),dim=c(5,2,2))
attr(B, "scale") <- "liter"
h5write(B, "myhdf5file.h5","foo/B")
C = matrix(paste(LETTERS[1:10],LETTERS[11:20], collapse=""),
  nr=2,nc=5)
h5write(C, "myhdf5file.h5","foo/foobaa/C")
df = data.frame(1L:5L,seq(0,1,length.out=5),
  c("ab","cde","fghi","a","s"), stringsAsFactors=FALSE)
h5write(df, "myhdf5file.h5","df")
h5ls("myhdf5file.h5")
D = h5read("myhdf5file.h5","foo/A")
E = h5read("myhdf5file.h5","foo/B")
F = h5read("myhdf5file.h5","foo/foobaa/C")
G = h5read("myhdf5file.h5","df")
```
