---
title: "Sparse data manipulation"
author: "Michael Love"
date: 11/7/2018
output: html_document
---

In this last lecture note on large data manipulation in R, we change
tactics a bit. Previously we discussed fast reading and subsetting
with *data.table*, and the advantages of *SQLite* vs *HDF5* for
storing collections of tables in a single file, and then working with
these in R using the *RSQLite* and *rhdf5* libraries. Here we discuss
an alternative approach for dealing with large arrays in which many
of the features are equal to zero. There are special classes and
methods in R that allow us to work with such data in a memory and
computationally efficient manner. These data are typically referred to
as *sparse* data, in that the non-zero elements of the array are
sparse. We will focus in this lecture note on the classes in the
*Matrix* package, and some functionality in the *glmnet* package for
fitting regularized linear or generalized linear models to sparse
feature matrices.

# Representing sparse matrices

Let's dive right into representing sparse matrices. Here we have a
large-ish matrix wherein the non-zero elements make up only ~5% of the
total:

```{r}
m <- matrix(rbinom(1e6, 1, .05), ncol=1e3)
m[1:5,1:5]
sum(m)
prod(dim(m))
```

This matrix takes up about 4 Mb in memory:

```{r}
print(object.size(m), units="Mb")
```

That's actually not so big that we encounter problems on a laptop
computer, but if we multiply either or both of the dimensions by a
factor of 1000, we will start to hit a limit in terms of working with
the matrix.

Let's get a sense of how much space we save if we represent this as a
sparse matrix.

```{r}
library(Matrix)
mm <- Matrix(m, sparse=TRUE)
mm[1:5,1:5]
sum(mm)
print(object.size(mm), units="Mb")
as.numeric(object.size(m)/object.size(mm))
```

The sparse version takes up less than 1/6 of the space of the *dense*
version. 

# How we actually construct sparse matrices

This coercion above, of a dense matrix into a sparse one doesn't make
any sense: we would never want to first build the memory-intensive
dense version of the matrix and then convert down to the sparse
version. Instead, we would use the `sparseMatrix` function to build
the matrix by specifying only the non-zero elements, and where they
occur.

First look up the help page for `sparseMatrix`:

```{r eval=FALSE}
?sparseMatrix
```

The most common way to construct a sparse matrix would be to specify
`i`, `j`, and `x` (this last argument optional, if not included, the
values will be equal to 1).

```{r}
s <- sparseMatrix(i=c(1,3,5),j=c(1,2,3),x=c(4,5,6),dims=list(6,4))
s
```

This creates an object of type `dgCMatrix`. Take a look at the help
page for this class

```{r eval=FALSE}
?dgCMatrix-class
```

You can see that this class is *column-oriented* which means it should 
be faster to index columns of these objects than rows. Likewise,
if we had not specified `x`, it would also be column-oriented by
default, but instead it would be `ngCMatrix`. Let's do a
microbenchmark to see the degree of difference. For this example,
column indexing is about twice as fast.

```{r warning=FALSE}
library(microbenchmark)
n <- 1e3
nn <- 1e5
s <- sparseMatrix(i=sample(n,nn,TRUE),
                  j=sample(n,nn,TRUE),
                  dims=list(n,n))
microbenchmark(sum(s[,10]),sum(s[10,]))
```

# Manipulating sparse matrices

We can do many operations to sparse matrices using specialized
functions which are different than the ones defined for regular
matrices. These are described in `?dgCMatrix-class`, but some of the
important ones are `%*%`, `crossprod`, `tcrossprod`, `solve`, `qr`,
`lu`. Using these operations will preserve the sparsity of the object
(so keeping us under our memory budger), and will perform much faster
than coercion to dense would, if the matrices have a high degree of
sparsity. 

Note that some operations destroy the sparsity, such as adding 1, and
therefore must be avoided (in the case where the dense matrix would
not fit in memory):

```{r}
s[1:10,1:10] + 1
```

Other operations maintain the sparsity:

```{r}
s[1:10,1:10] * 2
```

We can also plot an `image`, which avoids creating the dense
matrix: 

```{r}
image(s[1:100,1:100])
```

# Use of sparse matrices in glmnet

A common use case of sparse matrices is in prediction of a target,
let's call `y`, using a high-dimensional, sparse matrix of features
`x`. We are often in situation that there are more features in `x`
than there are observations (rows of `x` and length of `y`). In this
case it may make sense to first try linear modeling of `y` on `x`,
and to use some combination of L1 and L2 regularization to stabilize
the regression. The *glmnet* package allows one to fit elastic net
models for such a problem, where the `x` matrix can be sparse, and it
builds off of the sparse matrices defined in the *Matrix*
package. Read over the help file for the main function:

```{r echo=FALSE}
library(glmnet)
```

```{r eval=FALSE}
library(glmnet)
?glmnet
```

We can mock up some simulated data to show the kinds of models that
can be fit using `glmnet`. Here we simulate 50 columns of `x` with a
coefficient of `1` and the rest of the columns of `x` are not used in
constructing `y`.

```{r}
n <- 1e3
nn <- 1e5
x <- sparseMatrix(i=sample(n,nn,TRUE),
                  j=sample(n,nn,TRUE),
                  dims=list(n,n))
beta <- rep(c(1,0),c(50,950))
y <- x %*% beta + rnorm(n,0,.25)
```

Running `glmnet` gives us our regularization path. Here we set
`alpha=1` which corresponds to only the L1 penalty (lasso). Plotting
the regularization path reveals a range of `lambda` where the 50
coefficients have been correctly identified (non-zero coefficients)
while the rest of the coefficients have been shrunk to 0.

Notably, for this lecture note, we never had to convert `x` into its
dense form, thereby allowing much higher dimensions than would be
possibly if `glmnet` only took dense matrices as input.

```{r}
fit <- glmnet(x, y, family="gaussian", alpha=1)
plot(fit)
```
