---
title: "HW 10 - advMCMC"
author: "Naim Rashid"
date: "2/20/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# MCMC extension of HW 6

We would like to simulate from the posterior distribution of parameter $\boldsymbol{\theta} = (\pi,\lambda)$ pertaining to the fishing dataset and zero-inflated poisson model described in HW 6, assuming $\pi$ has a Unif(0,1) prior, and $\lambda$ has a Gamma(2,2) prior (shape and scale = 2).  The joint posterior can be written as $f(\pi,\lambda | \boldsymbol{y}) \propto f(\boldsymbol{y} | \pi, \lambda)f(\pi,\lambda) = f(\boldsymbol{y} | \pi, \lambda)f(\pi)f(\lambda)$, where $f(\boldsymbol{y} | \pi,\lambda)$ is the likelihood give in HW 6 except with $\lambda$ unknown, $f(\pi)$ is the specified prior for $\pi$, and $f(\lambda)$ is the specified prior for $\lambda$.  

Implement a MH random walk procedure to sample from the joint posterior of $\boldsymbol{\theta} = (\pi,\lambda)$.  You do not necessarily need to do a change of variable for $\pi$ or $\lambda$, however for proposals that exceed the boundaries of the parameter space of either parameter, the posterior for the propsal should be set = 0 (MH ratio = 0).  You may want to consider a narrower random walk variance in such as setting as well. 

You may use the following code below to get started, using $M = 20000$, random seed (1), starting values ($\pi^{(0)} = 0.3$, $\lambda = 3$), and burn-in period (2000 iterations) for all implementations of the algorithm below. Report the posterior means for $\pi$ and $\lambda$, as well as diagnostics such as trace plots and autocorrelation plots.

```{r}
### HELPER FUNCTIONS

# log prior for lambda, fill in 
lplambda = function(lambda){
  
  ## start solution

  ## end solution
  
}

# log prior for pi, fill in 
lppi = function(pi){
  
  ## start solution

  ## end solution  
  
}

# bivariate RW proposal function
# hint: bivariate proposal same as generating two indep proposals here
h.sim = function(){
  
  ## start solution

  ## end solution  
  
}

# returns ll, or log f(y|lambda, pi)
# compute given y and ny from table
ll = function(y, ny, x){
  pi = x[1]
  lambda = x[2]
  
  ## start solution

  ## end solution
}

# MH ratio
# Hint; since h symmetric, proposal density cancels out of ratio
R = function(y, y_weight, x, xt){
  # x is the proposal, xt is current state
  # x[1],xt[1] pertain to pi, x[2],xt[2] pertain to lambda
  ## start solution

  ## end solution
}

```

Now start the main code for the sampler

```{r}
# set the seed
set.seed(1)

# data fro HW 6
y = 0:6
ny = c(3062, 587, 284, 103, 33, 4, 2)

# Set chain length
M = 20000

# initialize the chain vector (alpha, lambda)
x.rw.chain = matrix(0, M, 2)
colnames(x.rw.chain) = c("pi","lambda")

# Initialize chain with specified initial values
# alpha, lambda
x.rw.chain[1,] = c(0.3, 3) 

# now start chain
for(i in 1:(M-1)){
  
  # set the value at current iteration of the chain to variable xt
  xt = 
  
  # draw a proposal from the proposal density
  x = 
  
  # calculate MH ratio 
  r = 
  
  # Generate draw from bernoulli(p).
  keep = rbinom(1, 1, r)
  
  # if keep = 1, then set next iteration equal to then proposal
  if(keep == 1){
    x.rw.chain[i+1,] = x
  }else{
    # otherwise, carry over value from the current iteration
    x.rw.chain[i+1,] = xt
  }
}

## print posterior means and diagnostic plots.  Comment on convergence 


```





